---
title: 'Effective sample size for the Cox model: insights into individual prediction
  uncertainty'
authors:
- name: Toby Hackmann
  orcid: "0009-0006-0540-7231"
  affiliation:
  - ref: lumc
  roles: analysis, writing
  corresponding: true
- name: Doranne Thomassen
  affiliation:
  - ref: lumc
  roles: conceptualization
- name: Saskia le Cessie
  affiliation:
  - ref: lumc
  - ref: epi
- name: Hein Putter
  affiliation:
  - ref: lumc
- name: Liesbeth C de Wreede
  affiliation:
  - ref: lumc
- name: Ewout W Steyerberg
  affiliation:
  - ref: lumc
  - ref: julius
# output:
#   pdf_document: default
#   word_document: default
affiliations:
- id: lumc
  name: Department of Biomedical Data Sciences, Leiden University Medical Center,
    Leiden, the Netherlands
- id: epi
  name: Department of Clinical Epidemiology, Leiden University Medical Center, Leiden,
    the Netherlands
- id: julius
  name: Julius Center for Health Sciences and Primary Care, University Medical Center
    Utrecht, Utrecht, the Netherlands
abstract: "Prognostic models are important to provide predictions for survival probabilities of individual patients. We aimed to explore the uncertainty for the Cox proportional hazards model for prediction of survival probabilities at different time points during follow-up. We hereto expand the concept of effective sample size, which translates uncertainty into a number of ‘patients like you’ that a prediction is effectively based on. Effective sample size and its properties were investigated in a real data example of colon cancer patients. Simulation studies clarified differences between hypothesized and observed behavior, related to the variance caused by the baseline hazard and coefficient estimation. Surprisingly, effective sample size for predictions of specific patients early on in follow-up can be larger than the total development sample size. While this can be considered misleading in communication with patients, we see value in using effective sample size for the investigation of individual uncertainty during model development.
In conclusion, uncertainty of predictions from the Cox model has unintuitive aspects which limits the suitability of measures such as effective sample size for communication with patients for relatively early follow-up times. More research is needed on the role of effective sample size for at late follow-up times."
number-sections: true
bibliography: references.bib
csl: american-medical-association.csl
---


## Introduction
Prognostic models with survival outcomes are abundant in disease areas such as oncology and cardiology. Examples are the Melanoma Institute Australia models which provide survival predictions for multiple stages of melanoma,[@pires2022; @varey2024] the PREDICT model breast cancer treatment decision support,[@predict] and the QRISK3 algorithm predicting the 10-year risk of cardiovascular disease.[@qrisk] Proper implementation of such prediction algorithms holds the promise to improve shared decision-making.[@4Dpicture] 

Uncertainty is an important element of predictions from prognostic models. It is unclear if and how uncertainty in the predicted survival probability should be communicated. Some risk calculators make no mention of prediction uncertainty,[@qrisk; @pires2022] while others show 95% confidence intervals around predicted survival probabilities[@varey2024] or treatment benefit.[@predict]

During model development, it is often considered that a model should have some minimal level of certainty in its predictions. For example, we may require a minimum number of events of 100 in order to estimate the overall event rate with some certainty.[@steyerbergbook;@harrellblog] Moreover, larger model complexity implies a need for larger numbers, as reflected in rules of thumb for the number of events per variable (EPV) such as 10 to 15 EPV for proportional hazards regression.[@peduzzi1995; @harrellbook] Such general rules have received criticism[@vansmeden2016] and novel methods calculate sample size or EPV considering more aspects of the prediction model.[@riley2019] More recently, focus has shifted to the stability of models and the individual uncertainty in predictions.[@altman1989; @shen2016; @riley2023; @riley2023b; @barrenada2025; @riley2025]

Effective sample size has been suggested as a promising summary measure of individual uncertainty in generalized linear models and machine learning models.[@thomassen2024;@thomassen2025] Effective sample size for a prediction for a patient can be interpreted as the number of patients with the same covariate values that the prediction was effectively based on. For analysis of survival data, the concept of effective sample size was presented already by Kaplan and Meier.[@kaplanmeier1958] We recently illustrated that effective sample size provides a valuable and readily interpretable measure of uncertainty for survival curves in the presence of censoring.[@hackmann2025] Effective sample size needs further extension to survival predictions based on combinations of prognostic factors, as commonly provided by a Cox proportional hazards model in medical research. 

In this study, we aim to extend previously developed approaches[@thomassen2025;@hackmann2025] for effective sample size estimation to the Cox model. We hypothesize that effective sample size may be valuable for expressing individual uncertainty in predictions based on multiple prognostic factors, with appropriate handling of censored observations as for Kaplan-Meier curves. For effective sample size to be an intuitive measure of uncertainty, we have two requirements that we will investigate:
1.	Natural behavior: Effective sample size reflects the size of a sample with complete follow-up, accommodating censoring 
2.	Reflection of exceptionality: Effective sample size reflects the exceptionality of a patient profile within the development data
This paper continues with of a description of calculating the effective sample size based on the variance of predictions of the Cox model (section 2), followed by application of effective sample size calculations to Cox models in an illustrative data set (section 3). We explore effective sample size in more depth in simulations (section 4) and end with a discussion (section 5), where we reflect on its potential use in communication and model development.


## Effective sample size for the Cox model {#sec-data-methods}
### Effective sample size
We consider a survival prediction for a specific new patient named ('Sam') at a predetermined time $t$ using a Cox model. This is a personalized survival prediction $\hat S(t | \mathbf{x}_{Sam}) = \hat S_{Sam}(t)$, based on their covariate values $\mathbf{x}_{Sam}$. We take as a reference the situation where we have a number $N_*$ of patients that all have the same covariate values $\mathbf{x}_{Sam}$ as Sam. In this virtual sample, we can observe the mean outcome $\bar S_*(t)$ at time $t$. With $\bar S_*(t) = \hat S_{Sam}(t)$, we can calculate the number $N_*$ as the number of patients required to obtain the same certainty in our estimation of $\bar S_*(t)$, as the model estimation of $\hat S_{Sam}(t)$. This number $N_*$ is the effective sample size.

If we assume that all patients with the same covariate values have the same variance, we can estimate $N_*$ using:
$$
N_* := \frac{V(\bar S_*(t))}{V(\hat S_{Sam}(t))},
$${#eq-n_eff}
where $V(\bar S_*(t)) = \bar S_*(t)\left(1-\bar S_*(t)\right)$ is the binomial variance and $V(\hat S_{Sam}(t))$ is the variance of the Cox model estimate for Sam. 

### Variance
For the variance of the Cox model, different estimation procedures of the variance can be used. We will use a standard definition for the Cox model variance based on the delta method [@tsiatis1981; @andersen1993], which is implemented in frequently used software for survival analysis [@package-survival; @mstate2011].  The variance of a Cox model prediction for Sam is:

$$
V(\hat S_{Sam}(t)) = \hat V\left( \hat S(t | \mathbf{x}_{Sam})\right)\biggr\rvert_{\beta} + \left( \frac{\partial\hat S}{\partial\beta}\right)\Biggr\rvert^\intercal_{\beta = \hat \beta} \hat V(\hat \beta) \left( \frac{\partial\hat S}{\partial\beta}\right)\Biggr\rvert_{\beta = \hat \beta},
$${#eq-cox_var}
where $\hat V\left( \hat S(t| \mathbf{x}_{new})\right)\biggr\rvert_{\beta}$ is the variance of the baseline hazard under correct coefficient estimation and $\left( \frac{\partial\hat S}{\partial\beta}\right)\Biggr\rvert^\intercal_{\beta = \hat \beta} \hat V(\hat \beta) \left( \frac{\partial\hat S}{\partial\beta}\right)\Biggr\rvert_{\beta = \hat \beta}$ is the variance caused by estimation of the coefficients. The variance of the baseline hazard can be estimated using:
$$
\hat{S}^2(t| \mathbf{x}_{new}) \mathrm{exp}(2\hat{\mathbf{\beta}}^\top\mathbf{x}_{Sam}) \left( \sum_{i=1}^l \frac{1}{\left\{ \sum_{j\in R_i} \mathrm{exp}(\hat{\mathbf{\beta}}^\top \mathbf{x}_{j}) \right\}^2} \right),
$${#eq-cox_var_basehaz}
which is the Aalen formula for the variance of the baseline survival function based on the risk-weighted number at risk, multiplied by the squared relative risk of our new patient Sam. 

The uncertainty from the estimation of the coefficient depends on the variance of the coefficients and the gradient of the survival function w.r.t. the coefficients. The variance of the coefficients $\hat V(\hat\beta)$, is the inverse Fisher information matrix $\mathcal{I}(\beta)^{-1}$. Finally, we can estimate the gradient of the survival function with respect to the coefficient by:
$$
\frac{\partial\hat S(t| \mathbf{x}_{new})}{\partial\beta} = \mathrm{exp}(\hat{\mathbf{\beta}}^\top \mathbf{x}_{new}) \mathbf{w},
$${#eq-cox_var_grad}
where $\mathbf{w}$ is a vector of partial derivatives, with $\mathbf{w}_p$ the partial derivative of the survival function in the direction of $\beta_p$, with:

$$
w_p = \sum_{i=1}^l \frac{1}{\sum_{j\in R_i}\mathrm{exp}(\hat{\mathbf{\beta}}^\top \mathbf{x}_{j})} \left\{ \frac{\sum_{j\in R_i} x_{j,p}\mathrm{exp}(\hat\beta^\top \mathbf{x}_j)}{ \sum_{j\in R_i} \mathrm{exp}(\hat{\mathbf{\beta}}^\top \mathbf{x}_{j})} - x_{new,p} \right\}.
$${#eq-cox_var_wp}
We can see in the definition of $\mathbf{w}_p$ that the variance, and therefore effective sample size, depends on the difference between our patient's new covariate $x_{Sam, p}$ and the distribution of $x_p$ in the risk set.

Note that other methods exist for calculating the variance, such as the Von Mises expansion[@vandervaart1998; @riskRegression], or by using the infinitesimal jackknife.[@survival-vignette] Any method works for estimating effective sample size using @eq-n_eff, however the estimate may be different, because of differences in estimated variance.

### Software
All analyses were conducted in `R`[@R], version 4.5.1, using the packages `survival`[@package-survival] and `mstate`.[@mstate2011] Simulations were run using the `simsurv` package.[@package-simsurv] Code used for this article, including functions that can calculate the effective sample size for a `survfit.coxph` object in `R` can be found on GitHub: xxxxx.

## Application to colon cancer data
### Data description
We performed analyses in the `colon` data set as available within the `survival` package[@package-survival] in `R` . This data set contains patients enrolled in a trial of adjuvant levimasole or levimasole+5-fluorouracil compared to observation only in 929 patients.[@laurie1989; @moertel1995] Mortality was the primary outcome, which was observed in 452 of 929 patients. Four baseline prognostic factors were considered in this study: with treatment as randomized (‘rx’); adhesion of the tumor (‘adhere') as categorical variables; and age (‘age’) and number of positive lymph nodes (‘nodes’) as continuous variables (Figure S1).

### Illustration with one predictor
We compared Cox models with a single predictor to Kaplan-Meier curves for the two categorical variables. Effective sample sizes for Cox model predictions were much higher than those of stratified Kaplan-Meier curves for early time points. Effective sample size was higher than the total development sample size for early survival estimates of patients in the Lev+5FU treatment arm and those in the No adhesion group. This difference in effective sample size decreased over follow-up time with increasing number of events (@fig-category). The higher effective sample for the Cox predictions matched with narrower confidence intervals around predicted survival curves compared to the Kaplan-Meier curves for these groups, especially early on in follow-up (Figure S2).

For Cox models including a continuous covariate as single predictor, effective sample size depended on covariate values. Patients with a common (median) age, the effective sample size was close to the total number of patients in the population until censoring started to impact effective sample size (@fig-continuous). For patients with uncommon (75th percentile) or rare (99th percentile) ages, the effective sample size was very low during early follow-up. For all three example patients, their survival predictions (dotted lines) were very similar, since the prognostic effect of age was very small. So, the uncertainty was large for predictions for patients with relatively rare ages, despite minor prognostic impact of age.

For the number of nodes, the common value was 2 positive nodes (median), while 5 was uncommon (75th percentile) and 17 was rare (99th percentile). Patients with 2 nodes had an effective sample size of more than 1000 early on in follow-up, which was higher than the total sample size of 929. Remarkably, for a patient with 17 nodes, the effective sample size of their prediction increased over time as their survival probability approached 0 (@fig-continuous).


```{r}
#| echo: FALSE
#| output: FALSE
#| message: FALSE
#| warning: FALSE
#| label: fig-category
#| fig-width: 10
#| fig-height: 6.2
#| fig-cap: "Effective sample size as a function of time for Cox model predictions based on the categorical variables `adhere` and `rx` in the `colon` dataset. Early on in follow-up, effective sample size for the Cox model survival curves is much higher than the Kaplan-Meier effective sample size and can even be higher than the development sample size. Effective sample size from Cox model predictions become similar to the effective sample sizes from stratified Kaplan-Meier curves at the end of follow-up."


source("code/compile.R")
colon_df$time <- colon_df$time/365.241

# Fit all the models
cox_adhere <- coxph( Surv(time, status) ~ adhere, data = colon_df)
cox_rx <- coxph( Surv(time, status) ~ rx, data = colon_df)

# Predict the effective sample size
km_adhere_0 <- survfit_n( survfit(Surv(time, status) ~ 1, data = colon_df[colon_df$adhere == 0,]) )
km_adhere_1 <- survfit_n( survfit(Surv(time, status) ~ 1, data = colon_df[colon_df$adhere == 1,]) )
km_rx_0 <- survfit_n( survfit(Surv(time, status) ~ 1, data = colon_df[colon_df$rx == "Obs",]) )
km_rx_1 <- survfit_n( survfit(Surv(time, status) ~ 1, data = colon_df[colon_df$rx == "Lev",]) )
km_rx_2 <- survfit_n( survfit(Surv(time, status) ~ 1, data = colon_df[colon_df$rx == "Lev+5FU",]) )
cox_adhere_0 <- survfit_n( survfit(cox_adhere, newdata=data.frame(adhere=0)), cox=cox_adhere)
cox_adhere_1 <- survfit_n( survfit(cox_adhere, newdata=data.frame(adhere=1)), cox=cox_adhere)
cox_rx_0 <- survfit_n( survfit(cox_rx, newdata=data.frame(rx="Obs")), cox=cox_rx)
cox_rx_1 <- survfit_n( survfit(cox_rx, newdata=data.frame(rx="Lev")), cox=cox_rx)
cox_rx_2 <- survfit_n( survfit(cox_rx, newdata=data.frame(rx="Lev+5FU")), cox=cox_rx)

color_values <- c("Obs"= "#2E7691", "Lev" = "#C2666B", "No Adhesion"= "#37293F", "Lev+5FU" = "#c6aa2c", "Adhesion" = "#9f84af")
linetype_values <- c("Cox" = "solid", "Kaplan-Meier" = "dashed")

# Create the plots
plot1 <- ggplot() +
  geom_line(aes(x = cox_rx_0$time, y = cox_rx_0$surv*100, color = "Obs", linetype = "Cox"), lwd = 1.3) +
  geom_line(aes(x = cox_rx_1$time, y = cox_rx_1$surv*100, color = "Lev", linetype = "Cox"), lwd = 1.3) +
  geom_line(aes(x = cox_rx_2$time, y = cox_rx_2$surv*100, color = "Lev+5FU", linetype = "Cox"), lwd = 1.3) +
  geom_line(aes(x = km_rx_0$time, y = km_rx_0$surv*100, color = "Obs", linetype = "Kaplan-Meier"), lwd = 1.3) +
  geom_line(aes(x = km_rx_1$time, y = km_rx_1$surv*100, color = "Lev", linetype = "Kaplan-Meier"), lwd = 1.3) +
  geom_line(aes(x = km_rx_2$time, y = km_rx_2$surv*100, color = "Lev+5FU", linetype = "Kaplan-Meier"), lwd = 1.3) +
  scale_color_manual(values = color_values) +
  scale_linetype_manual(values = linetype_values) +
  theme_minimal() +
  scale_y_continuous(name = "Predicted survival (%)", limits = c(0, 100)) +
  scale_x_continuous(name = NULL, labels = NULL,
                     breaks = seq(0, max(colon_df$time), by = 2)) +
  ggtitle("Model based on treatment") +
  theme(legend.position = "none") # Remove legend from individual plot

plot2 <- ggplot() +
  geom_line(aes(x = cox_rx_0$time, y = cox_rx_0$n.eff, color = "Obs", linetype = "Cox"), lwd = 1.3) +
  geom_line(aes(x = cox_rx_1$time, y = cox_rx_1$n.eff, color = "Lev", linetype = "Cox"), lwd = 1.3) +
  geom_line(aes(x = cox_rx_2$time, y = cox_rx_2$n.eff, color = "Lev+5FU", linetype = "Cox"), lwd = 1.3) +
  geom_line(aes(x = km_rx_0$time, y = km_rx_0$n.eff, color = "Obs", linetype = "Kaplan-Meier"), lwd = 1.3) +
  geom_line(aes(x = km_rx_1$time, y = km_rx_1$n.eff, color = "Lev", linetype = "Kaplan-Meier"), lwd = 1.3) +
  geom_line(aes(x = km_rx_2$time, y = km_rx_2$n.eff, color = "Lev+5FU", linetype = "Kaplan-Meier"), lwd = 1.3) +
  geom_hline(yintercept = 929, linetype = "dotted", color = "black", linewidth = 0.6) +
  scale_color_manual(values = color_values) +
  scale_linetype_manual(values = linetype_values) +
  theme_minimal() +
  scale_y_continuous(name = "Effective Sample Size", limits = c(0, 1000)) +
  scale_x_continuous(name = "Time (years)",
                     breaks = seq(0, max(colon_df$time), by = 2)) +
  theme(legend.position = "none")  # Remove legend from individual plot

plot3 <- ggplot() +
  geom_line(aes(x = cox_adhere_0$time, y = cox_adhere_0$surv*100, color = "No Adhesion", linetype = "Cox"), lwd = 1.3) +
  geom_line(aes(x = cox_adhere_1$time, y = cox_adhere_1$surv*100, color = "Adhesion", linetype = "Cox"), lwd = 1.3) +
  geom_line(aes(x = km_adhere_0$time, y = km_adhere_0$surv*100, color = "No Adhesion", linetype = "Kaplan-Meier"), lwd = 1.3) +
  geom_line(aes(x = km_adhere_1$time, y = km_adhere_1$surv*100, color = "Adhesion", linetype = "Kaplan-Meier"), lwd = 1.3) +
  scale_color_manual(values = color_values) +
  scale_linetype_manual(values = linetype_values) +
  theme_minimal() +
  scale_y_continuous(name = NULL, labels = NULL, limits = c(0, 100)) +  # Remove y-axis name and tick labels
  scale_x_continuous(name = NULL, labels = NULL,
                     breaks = seq(0, max(colon_df$time), by = 2)) +
  ggtitle("Model based on adhesion") +
  theme(legend.position = "none")  # Remove legend from individual plot

plot4 <- ggplot() +
  geom_line(aes(x = cox_adhere_0$time, y = cox_adhere_0$n.eff, color = "No Adhesion", linetype = "Cox"), lwd = 1.3) +
  geom_line(aes(x = cox_adhere_1$time, y = cox_adhere_1$n.eff, color = "Adhesion", linetype = "Cox"), lwd = 1.3) +
  geom_line(aes(x = km_adhere_0$time, y = km_adhere_0$n.eff, color = "No Adhesion", linetype = "Kaplan-Meier"), lwd = 1.3) +
  geom_line(aes(x = km_adhere_1$time, y = km_adhere_1$n.eff, color = "Adhesion", linetype = "Kaplan-Meier"), lwd = 1.3) +
  geom_hline(yintercept = 929, linetype = "dotted", color = "black", linewidth = 0.6) +
  scale_color_manual(values = color_values) +
  scale_linetype_manual(values = linetype_values) +
  theme_minimal() +
  scale_y_continuous(name = NULL, labels = NULL, limits = c(0, 1000)) +  # Remove y-axis name and tick labels
  scale_x_continuous(name = "Time (years)",
                     breaks = seq(0, max(colon_df$time), by = 2)) +
  theme(legend.position = "none")  # Remove legend from individual plot

# Combine the plots using patchwork and add a single legend on the right
combined_plot <- (plot1 / plot2 | plot3 / plot4) +
  plot_layout(guides = "collect") &
  theme(legend.position = "right")

ggsave("images/cox_ess_categories.svg", width = 13, height = 7, dpi = 300)
# Print the combined plot
print(combined_plot)


color_values <- c("Obs"= "#2E7691", "Kaplan-Meier" = "#C2666B", "Cox"= "#37293F", "Lev+5FU" = "#c6aa2c", "Adhesion" = "#9f84af")
linetype_values <- c("Prediction" = "solid", "95% CI" = "dashed")
plotx <- ggplot() +
  # Cox curves
  geom_line(aes(x = cox_rx_2$time, y = cox_rx_2$surv, color = "Cox", linetype = "Prediction"),
            lwd = 1.3, alpha = 0.5) +
  geom_line(aes(x = cox_rx_2$time, y = cox_rx_2$upper, color = "Cox", linetype = "95% CI"),
            lwd = 1.3, alpha = 0.5) +
  geom_line(aes(x = cox_rx_2$time, y = cox_rx_2$lower, color = "Cox", linetype = "95% CI"),
            lwd = 1.3, alpha = 0.5) +
  
  # Kaplan–Meier curves
  geom_line(aes(x = km_rx_2$time, y = km_rx_2$surv, color = "Kaplan-Meier", linetype = "Prediction"),
            lwd = 1.3, alpha = 0.5) +
  geom_line(aes(x = km_rx_2$time, y = km_rx_2$upper, color = "Kaplan-Meier", linetype = "95% CI"),
            lwd = 1.3, alpha = 0.5) +
  geom_line(aes(x = km_rx_2$time, y = km_rx_2$lower, color = "Kaplan-Meier", linetype = "95% CI"),
            lwd = 1.3, alpha = 0.5) +
  
  # Double arrow for Kaplan–Meier at t = 480
  geom_segment(
    {
      i <- which.max(ifelse(km_rx_2$time < 480, km_rx_2$time, -Inf))
      aes(x = km_rx_2$time[i], xend = km_rx_2$time[i],
          y = km_rx_2$lower[i], yend = km_rx_2$upper[i])
    },
    color = color_values["Kaplan-Meier"],
    linewidth = 1.2,
    arrow = arrow(length = unit(0.15, "cm"), ends = "both", type = "closed")
  ) +
  
  # Double arrow for Cox at t = 520
  geom_segment(
    {
      j <- which.max(ifelse(cox_rx_2$time < 520, cox_rx_2$time, -Inf))
      aes(x = cox_rx_2$time[j], xend = cox_rx_2$time[j],
          y = cox_rx_2$lower[j], yend = cox_rx_2$upper[j])
    },
    color = color_values["Cox"],
    linewidth = 1.2,
    arrow = arrow(length = unit(0.15, "cm"), ends = "both", type = "closed")
  ) +
  
  scale_color_manual(values = color_values) +
  scale_linetype_manual(values = linetype_values) +
  theme_minimal() + 
  scale_y_continuous(name = "Survival probability") +
  scale_x_continuous(name = "Time (years)",
                     breaks = seq(0, max(colon_df$time), by = 2)) +
  labs(title = "Survival curve of a patient with Lev+5FU-treatment from Cox and Kaplan-Meier")


#plotx
ggsave("images/compare_ci.svg", width = 13, height = 7, dpi = 300)
#save(plotx, file = "images/fig-compare_ci.RData" )
#rm(list = ls())

```


```{r}
#| echo: FALSE
#| output: FALSE
#| message: FALSE
#| warning: FALSE
#| label: fig-continuous
#| fig-width: 10 
#| fig-height: 6.2
#| fig-cap: "Effective sample size as a function of time for Cox model predictions based on the continuous variables `age` and `nodes` in the `colon` dataset for a common value in the data (median), uncommon (75% percentile) and rare vairable (99% for age, 99.5% for nodes). The top row contains the plots for models based on age and the bottom row those based on nodes. On the left-hand-side are the Cox model survival predictions for each of the patients and on the right-hand-side are the effective sample sizes."

source("code/compile.R")
colon_df$time <- colon_df$time/365.241

# Calculate percentiles for age
age_percentiles <- quantile(colon_df$age, probs = c(0.5, 0.75, 0.99), na.rm = TRUE)
nodes_percentiles <- round(quantile(colon_df$nodes, probs = c(0.5, 0.75, 0.99), na.rm = TRUE))
nodes_percentiles["99%"] <- 20  # Manually set to 25 as it's more extreme than the 99th percentile

# Fit Cox models for age and nodes
cox_age <- coxph(Surv(time, status) ~ age, data = colon_df)
cox_nodes <- coxph(Surv(time, status) ~ nodes, data = colon_df)

# Predict effective sample size for age percentiles
cox_age_median <- survfit_n(survfit(cox_age, newdata = data.frame(age = age_percentiles["50%"])), cox = cox_age)
cox_age_75 <- survfit_n(survfit(cox_age, newdata = data.frame(age = age_percentiles["75%"])), cox = cox_age)
cox_age_99 <- survfit_n(survfit(cox_age, newdata = data.frame(age = age_percentiles["99%"])), cox = cox_age)

# Predict survival probabilities for age percentiles
surv_age_median <- survfit(cox_age, newdata = data.frame(age = age_percentiles["50%"]))
surv_age_75 <- survfit(cox_age, newdata = data.frame(age = age_percentiles["75%"]))
surv_age_99 <- survfit(cox_age, newdata = data.frame(age = age_percentiles["99%"]))

# Predict effective sample size for nodes with specific values
cox_nodes_2 <- survfit_n(survfit(cox_nodes, newdata = data.frame(nodes = nodes_percentiles["50%"])), cox = cox_nodes)
cox_nodes_5 <- survfit_n(survfit(cox_nodes, newdata = data.frame(nodes = nodes_percentiles["75%"])), cox = cox_nodes)
cox_nodes_25 <- survfit_n(survfit(cox_nodes, newdata = data.frame(nodes = nodes_percentiles["99%"])), cox = cox_nodes)

# Predict survival probabilities for nodes with specific values
surv_nodes_2 <- survfit(cox_nodes, newdata = data.frame(nodes = 2))
surv_nodes_5 <- survfit(cox_nodes, newdata = data.frame(nodes = 5))
surv_nodes_25 <- survfit(cox_nodes, newdata = data.frame(nodes = 25))

# Define color values
color_values <- c("Common" = "#37293F", "Uncommon" = "#2E7691", "Rare" = "#C2666B")
linetype_values <- c("Effective Sample Size" = "solid", "Survival Probability" = "dotted")

# Create the plots for age
plot1 <- ggplot() +
  geom_line(aes(x = surv_age_median$time, y = surv_age_median$surv*100, color = "Common"), lwd = 1.3) +
  geom_line(aes(x = surv_age_75$time, y = surv_age_75$surv*100, color = "Uncommon"), lwd = 1.3) +
  geom_line(aes(x = surv_age_99$time, y = surv_age_99$surv*100, color = "Rare"), lwd = 1.3) +
  scale_color_manual(values = color_values) +
  scale_linetype_manual(values = linetype_values) +
  theme_minimal() +
  scale_y_continuous(
    name = "Predicted survival (%)",
    limits = c(0, 100)
  ) +
  scale_x_continuous(labels = NULL, name = NULL,
                     breaks = seq(0, max(colon_df$time), by = 2)) +
  ggtitle("Cox model based on age") +
  theme(legend.position = "none")  # Remove legend from individual plot

# Create the plots for nodes
plot2 <- ggplot() +
  geom_line(aes(x = surv_nodes_2$time, y = surv_nodes_2$surv*100, color = "Common"), lwd = 1.3) +
  geom_line(aes(x = surv_nodes_5$time, y = surv_nodes_5$surv*100, color = "Uncommon"), lwd = 1.3) +
  geom_line(aes(x = surv_nodes_25$time, y = surv_nodes_25$surv*100, color = "Rare"), lwd = 1.3) +
  scale_color_manual(values = color_values) +
  scale_linetype_manual(values = linetype_values) +
  theme_minimal() +
  scale_y_continuous(
    limits = c(0, 100),
    labels = NULL,
    name = NULL
  ) +  # Remove y-axis name and tick labels
  scale_x_continuous(labels = NULL, name = NULL,
                     breaks = seq(0, max(colon_df$time), by = 2)) +
  ggtitle( "Cox model based on number of nodes") +
  theme(legend.position = "none")  # Remove legend from individual plot

plot3 <- ggplot() +
  geom_line(aes(x = cox_age_median$time, y = cox_age_median$n.eff, color = "Common"), lwd = 1.3) +
  geom_line(aes(x = cox_age_75$time, y = cox_age_75$n.eff, color = "Uncommon"), lwd = 1.3) +
  geom_line(aes(x = cox_age_99$time, y = cox_age_99$n.eff, color = "Rare"), lwd = 1.3) +
  geom_hline(yintercept = 929, linetype = "dotted", color = "black", linewidth = 0.6) +
  scale_color_manual(values = color_values) +
  scale_linetype_manual(values = linetype_values) +
  theme_minimal() +
  scale_y_continuous(
    name = "Effective Sample Size",
    limits = c(0, 1000)
  ) +
  scale_x_continuous(name = "Time (years)",
                     breaks = seq(0, max(colon_df$time), by = 2)) +
  theme(legend.position = "none")  # Remove legend from individual plot

# Create the plots for nodes
plot4 <- ggplot() +
  geom_line(aes(x = cox_nodes_2$time, y = cox_nodes_2$n.eff, color = "Common"), lwd = 1.3) +
  geom_line(aes(x = cox_nodes_5$time, y = cox_nodes_5$n.eff, color = "Uncommon"), lwd = 1.3) +
  geom_line(aes(x = cox_nodes_25$time, y = cox_nodes_25$n.eff, color = "Rare"), lwd = 1.3) +
  geom_hline(yintercept = 929, linetype = "dotted", color = "black", linewidth = 0.6) +
  scale_color_manual(values = color_values) +
  scale_linetype_manual(values = linetype_values) +
  theme_minimal() +
  scale_y_continuous(
    limits = c(0, 1000),
    labels = NULL,
    name = NULL
  ) +  # Remove y-axis name and tick labels
  scale_x_continuous(name = "Time (years)",
                     breaks = seq(0, max(colon_df$time), by = 2)) +
  theme(legend.position = "none")  # Remove legend from individual plot


# Combine the plots using patchwork and add a single legend on the right
combined_plot <- (plot1 / plot3 | plot2 / plot4) +
  plot_layout(guides = "collect") &
  theme(legend.position = "right")

# Print the combined plot
print(combined_plot)
ggsave("images/cox_ess_continuous.svg", width = 13, height = 7, dpi = 300)

rm( list = ls() )
```


### Effective sample size in multivariable Cox models
A survival prediction from a Cox model is often presented as the survival probability at specific times points $t$. It may not be relevant that effective sample size can reach unintuitive values higher than the development sample size at early follow-up times, if it is intuitively interpretable at the time point of interest. We fit four Cox models, each time adding a predictor: (A) Surv(time, status) ~ rx; (B) Surv(time, status) ~ rx + adhere; (C) Surv(time, status) ~ rx + adhere + nodes; (D) Surv(time, status) ~ rx + adhere + nodes + age. 

```{r}
#| echo: FALSE
#| output: FALSE
#| eval: FALSE
#| message: FALSE
#| warning: FALSE
#| fig-width: 10
#| label: fig-distribution-dev
#| fig-cap: "Distribution of effective sample size for the 5-year survival prediction in the data as a function of the predicted 5-year survival probability. The first model includes only treatment. The second model adds adhesion, the third model adds nodes and the fourth model adds age It can be seen that the effective sample size does not decrease for all patients with increasing variables, but it does for patients with rare covariate values."

source("code/compile.R")

year <- 5
var1 <- "rx"; var2 <- "rx + adhere"; var3 <- "rx + adhere + nodes"; var4 <- "rx + adhere + nodes + age"
leg1 <- "Tx"; leg2 <- "Tx + Adh"; leg3 <- "Tx + Adh + Nodes"; leg4 <- "Tx + Adh + Nodes + Age"


# Single impute the nodes data
colon_df$nodes <- mice::complete(mice::mice(colon_df, m = 1, print = FALSE))$nodes

# Fit all cox models
cox_model1 <- coxph(as.formula(paste0( "Surv( time, status) ~ ", var1)), data = colon_df)
cox_model2 <- coxph(as.formula(paste0( "Surv( time, status) ~ ", var2)), data = colon_df)
cox_model3 <- coxph(as.formula(paste0( "Surv( time, status) ~ ", var3)), data = colon_df)
cox_model4 <- coxph(as.formula(paste0( "Surv( time, status) ~ ", var4)), data = colon_df)

# Create prediction df
out_df1 = out_df2 = out_df3 = out_df4 <- colon_df |>
  select( rx, age, adhere, nodes )
out_df1$model <- paste0( "(A) Surv( time, status) ~ ", var1)
out_df2$model <- paste0( "(B) Surv( time, status) ~ ", var2)
out_df3$model <- paste0( "(C) Surv( time, status) ~ ", var3)
out_df4$model <- paste0( "(D) Surv( time, status) ~ ", var4)


# Create SF object for correct time
sf1 <- summary( survfit(cox_model1, newdata = out_df1), time = year*365.25 )
sf2 <- summary( survfit(cox_model2, newdata = out_df2), time = year*365.25 )
sf3 <- summary( survfit(cox_model3, newdata = out_df3), time = year*365.25 )
sf4 <- summary( survfit(cox_model4, newdata = out_df4), time = year*365.25 )

# Save survival probabilities
out_df1$surv <- c(sf1$surv)
out_df2$surv <- c(sf2$surv)
out_df3$surv <- c(sf3$surv)
out_df4$surv <- c(sf4$surv)

# Calculate the effective sample size for the 5-year prediction
out_df1$n.eff <- (1-out_df1$surv)*out_df1$surv/(c(sf1$std.err)^2)
out_df2$n.eff <- (1-out_df2$surv)*out_df2$surv/(c(sf2$std.err)^2)
out_df3$n.eff <- (1-out_df3$surv)*out_df3$surv/(c(sf3$std.err)^2)
out_df4$n.eff <- (1-out_df4$surv)*out_df4$surv/(c(sf4$std.err)^2)

out_df <- rbind(out_df1, out_df2, out_df3, out_df4)
out_df$model <- factor(out_df$model, levels = c(paste0( "(A) Surv( time, status) ~ ", var1), 
                                                paste0( "(B) Surv( time, status) ~ ", var2),  
                                                paste0( "(C) Surv( time, status) ~ ", var3), 
                                                paste0( "(D) Surv( time, status) ~ ", var4)))

# Create a scatter plot
colors <- c("#37293F", "#2E7691",  "#9f84af",  "#c6aa2c")
names(colors) <- c(paste0( "(A) Surv( time, status) ~ ", var1), 
                   paste0( "(B) Surv( time, status) ~ ", var2),  
                   paste0( "(C) Surv( time, status) ~ ", var3), 
                   paste0( "(D) Surv( time, status) ~ ", var4))

plot <- ggplot(out_df) +
  geom_point( aes(x = surv, y = n.eff, color = model), data = out_df, alpha = 0.2, size = 2) +
  facet_wrap( ~model ) +
  scale_x_continuous(name = "Predicted 5-Year Survival Probability", limits = c(0, 1)) +
  scale_y_continuous(name = "Effective Sample Size", limits = c(0, 500)) +
  #ggtitle("Changes distribution of Eff. N. between models") +
  labs( color = "Legend") +
  scale_color_manual(
    values = colors
  ) +
  #theme_minimal() +
  theme(legend.position = "none") + 
  guides(colour = guide_legend(override.aes = list(alpha = 1)))

# Print the plot
plot
save(plot, file = "images/fig-distribution.RData" )
ggsave("images/fig-distribution.svg", width = 13, height = 10, dpi = 300)
rm( list = ls())


LL_null <- cox_model2$loglik[1]   # null model loglik
n <- nrow(colon_df)

R2_CS <- function(LL_model) {
  1 - exp((2/n) * (LL_null - LL_model))
}

R2_base <- R2_CS(cox_model2$loglik[2])
R2_nodes <- R2_CS(cox_model3$loglik[2])
R2_age <- R2_CS(cox_model4$loglik[2])

delta_R2_nodes <- R2_nodes - R2_base
delta_R2_age <- R2_age - R2_nodes
print(paste0("Delta R^2 after addition of nodes is: ", round(delta_R2_nodes*100, 2), "%"))
print(paste0("Delta R^2 after addition of age is: ", round(delta_R2_age*100, 2), "%"))
```
```{r}
#| echo: FALSE
#| output: TRUE
#| message: FALSE
#| warning: FALSE
#| label: fig-distribution
#| fig-width: 10
#| fig-height: 6.2
#| fig-cap: "Distribution of effective sample size for the 5-year survival prediction in the data as a function of the predicted 5-year survival probability. The first model includes only treatment. The second model adds adhesion, the third model adds nodes and the fourth model adds age It can be seen that the effective sample size does not decrease for all patients with increasing variables, but it does for patients with rare covariate values."

load("images/fig-distribution.RData")

plot
```

The addition of adhesion to a model based only on treatment (‘rx’) changed the predicted probabilities, and decreased the effective sample size for the substantially for patients with lower survival predictions than in the treatment-only model. Similarly, the addition of nodes decreased the effective sample size for many patients, while also increasing the spread in predicted survival outcomes. In contrast, the addition of age to this model showed a minor increase in spread of the predictions, while there was much more heterogeneity in effective sample sizes. Indeed, age was a weak predictor (delta Cox-Snell $R^2$ 0.15\%) of survival outcome, whereas nodes was a much stronger predictor (delta Cox-Snell $R^2$ 7.66\%).

## Investigation of effective sample size in simulations
### Simulation setting
To understand when and why effective sample size can be higher than the development sample size, simulations were run without censoring and with one binary variable: ‘sex’. A hazard ratio of 2 for men compared to women was chosen for illustration. Sample size in the simulation was set to 10,000 (5,000 men and 5,000 women) to ensure that the hazard ratio could be accurately estimated between the two groups. Baseline hazard was set to a constant of 1, and maximum follow-up was set to 10 years.

### Simulation results
Cox model and Kaplan-Meier survival estimates were identical for both groups. The Kaplan-Meier effective sample sizes for both groups remained constant at 5,000. This is consistent with the number of observations in each of the groups because there was no censoring.[@hackmann2025] The Cox model effective sample size started at 15,000 for women (the group with lower hazard) and at 7,500 for men (the group with higher hazard). Additionally, the effective sample size for men increased after t=7.5.
To investigate both of these unexpected behaviors of effective sample size in more detail we separated the variance of the Cox model prediction into the variance caused by the estimation of the baseline hazard (first term in Equation 2) and the variance caused by the estimation of the coefficient (second term in Equation 2, Figure 4C and 4D ).


Early on in follow-up the women had a lower hazard and an effective sample size higher than the total sample size. This is attributable to the estimation of variance of the baseline hazard, since this component was the source of nearly all of the variance early on in follow-up (Figure 4A). The incremental increase in variance due to coefficient estimation for men was about the same as that of the baseline hazard after t=5. For women, it was about the same up to t=7.5, after which the increment of the coefficient variance became substantially smaller than that of the baseline hazard. This coincides with the inflection point where the percentage contribution of the baseline hazard began to increase again.



```{r}  
#| echo: FALSE
#| output: FALSE
#| eval: FALSE
#| message: FALSE
#| warning: FALSE
#| fig-width: 10
#| fig-height: 6.2
#| label: fig-sim-dev
#| fig-cap: "Simulation of survival data and effective sample sizes of survival estimates with no censoring. Effective sample size for Kaplan-Meier survival estimates is 5,000 all the time for both groups. For the Cox model predictions, the effective sample size begins at 7500 for men and 15000 for women. The explaination for this is that the weighted risk set is 7500 in the Cox model. When the first patient dies, a man has a 1/7500 chance of mortality, showing the effective sample size of 7500. A woman has a 0.5/7500 = 15000 risk of death, showing an assumed effective sample size of 15000."


source("code/compile.R")
require(surveillance)

# Seed for consistency
set.seed(1923)

scale = c(10)
shape = c(1) # shape 1 means exponential
maxtime = 100 # maxtime
hr = 2 # hazard ratio
n = 5000 # number per group

events <- simsurv( dist="weibull", 
                   gammas=shape, 
                   lambdas=scale[1]^(-shape), 
                   x=data.frame(tx = c(rep(0, n), rep(1, n))),
                   betas = c("tx" = log(hr)),
                   maxt=maxtime)
events$tx = c(rep(0, n), rep(1, n))
events$eventtime[8885] <- 2.997671  # handmatig ties weghalen
events$eventtime[6298] <- 5.668818  # werkt omdat seed vast staats
time <- sort(unique(events$eventtime))
time0 <- sort(unique(events[events$tx == 0, ]$eventtime))
time1 <- sort(unique(events[events$tx == 1, ]$eventtime))

# Fit the models
sf0 <- survfit(Surv(eventtime, status) ~ 1, data = events[events$tx == 0,])
sf1 <- survfit(Surv(eventtime, status) ~ 1, data = events[events$tx == 1,])
cox <- coxph(Surv(eventtime, status) ~ tx, data = events)

# Calculate effective N and stuff
sf_n0 <- survfit_n(sf0) |> sf_to_df(time = time0)
sf_n1 <- survfit_n(sf1) |> sf_to_df(time = time1)
cox_n0 <- survfit_n(survfit(cox, newdata = data.frame(tx = 0)), cox, coef=T, chaz=T)
cox_n1 <- survfit_n(survfit(cox, newdata = data.frame(tx = 1)), cox, coef=T, chaz=T)

cox_n0 <- sf_to_df(cox_n0, time = time)
cox_n1 <- sf_to_df(cox_n1, time = time)

colors <- c("1" = "#37293F", "3" = "#2E7691",  "2" = "#9f84af", "4" = "#c6aa2c")
linetype <- c("1" = "solid", "2" = "dashed")

# plot the effective sample sizes all in one plot
plot1 <- ggplot() +
  geom_line(aes(x = time, y = cox_n0$n.eff, color = "1"), size = 1.2) +
  geom_line(aes(x = time, y = cox_n1$n.eff, color = "2"), size = 1.2) +
  geom_line(aes(x = time1, y = sf_n0$n.eff, color = "3"), size = 1.2) +
  geom_line(aes(x = time1, y = sf_n1$n.eff, color = "4"), size = 1.2) +
  scale_x_continuous(name = NULL, labels = NULL, limits = c(0, 15)) +
  scale_y_continuous(name = "Effective Sample Size (N)", limits = c(0, 15000)) +
  #ggtitle("Effective Sample Size of Coxph and Survfit") +
  labs( color = "Legend", tag = "(B)") +
  scale_color_manual(
    values = colors,
    labels = c("Cox Women", "Cox Men", "KM Women", "KM Men")
  ) +
  theme_minimal() +
  theme(legend.position = "right") + 
  guides(colour = guide_legend(override.aes = list(alpha = 1)))


# plot the two kaplan-meiers overlapping in one plot and the cox survival predictions for the 0 an 1 treatment patients
plot2 <- ggplot() +
  geom_line(aes(x = time0, y = sf_n0$surv*100, color = "3"), size = 1.2) +
  geom_line(aes(x = time1, y = sf_n1$surv*100, color = "4"), size = 1.2) +
  geom_line(aes(x = time, y = cox_n0$surv*100, color = "1"), size = 1.2) +
  geom_line(aes(x = time, y = cox_n1$surv*100, color = "2"), size = 1.2) +
  scale_x_continuous(name = "Time", limits = c(0, 15)) +
  scale_y_continuous(name = "Predicted survival (%)", limits = c(0, 100)) +
  #ggtitle("Survival of Coxph and Survfit") +
  labs( color = "Legend", tag = "(A)") +
  scale_color_manual(
    values = colors,
    labels = c("Cox Women", "Cox Men", "KM Women", "KM Men")
  ) +
  theme_minimal() +
  theme(legend.position = "none") + 
  guides(colour = guide_legend(override.aes = list(alpha = 1))) 

time <- sort(unique(c(cox_n0$time, cox_n1$time)))

cox_incremental_hazvar0 <- (c(cox_n0$std.chaz, 0)-c(0, cox_n0$std.chaz))*cox_n0$surv
cox_incremental_hazvar1 <- (c(cox_n1$std.chaz, 0)-c(0, cox_n1$std.chaz))*cox_n0$surv
cox_incremental_coefvar0 <- (c(cox_n0$std.coef, 0)-c(0, cox_n0$std.coef))*cox_n0$surv
cox_incremental_coefvar1 <- (c(cox_n1$std.coef, 0)-c(0, cox_n1$std.coef))*cox_n0$surv
colors <- c("1" = "#37293F", "3" = "#2E7691",  "2" = "#9f84af", "4" = "#c6aa2c")
linetype <- c("1" = "solid", "2" = "dashed", "3" = "dotted")

scaleFactor = 0.00003/1

plot3 <- ggplot() +
  geom_line(aes(x = time, y = cox_n0$std.chaz^2, color = "1", lty = "1"), size = 1, lwd = 1.2) +
  geom_line(aes(x = time, y = cox_n1$std.chaz^2, color = "2", lty = "1"), size = 1, lwd = 1.2) +
  geom_line(aes(x = time, y = cox_n0$std.err^2, color = "1", lty = "2"), size = 1, lwd = 1.2) +
  geom_line(aes(x = time, y = cox_n1$std.err^2, color = "2", lty = "2"), size = 1, lwd = 1.2) +
  scale_x_continuous(name = "Time", limits = c(0, 15)) +
  scale_y_continuous(name = "Variance of Cox model", limits = c(0, 0.003)) +
  #ggtitle("Increment of standard error") +
  labs( color = "Legend", tag = "(D)") +
  scale_color_manual(
    values = colors,
    labels = c("Women", "Men")
  ) +
  scale_linetype_manual(
    values = linetype,
    labels = c( "Baseline hazard", "Total variance" ) ) +
  theme_minimal() +
  theme(legend.position = "none") + 
  guides(colour = guide_legend(override.aes = list(alpha = 1)))


plot4 <- ggplot() +
  geom_line(aes(x = time, y = cox_n0$std.chaz^2/cox_n0$std.err^2*100, color = "1", lty = "1"), size = 1, lwd = 1.2) +
  geom_line(aes(x = time, y = cox_n1$std.chaz^2/cox_n1$std.err^2*100, color = "2", lty = "1"), size = 1, lwd = 1.2) +
  scale_x_continuous(name = "Time", limits = c(0, 15)) +
  scale_y_continuous(name = "Percentage of variance from h0", limits = c(0, 100)) +
  
  #ggtitle("Proportion of variance from baseline hazard") +
  labs( color = "Legend", tag = "(C)") +
  scale_color_manual(
    values = colors,
    labels = c("Women", "Men")
  ) +
  scale_linetype_manual(
    values = linetype,
    labels = c( "Baseline hazard", "Coefficient" ) ) +
  theme_minimal() +  
  theme(legend.position = "right") + 
  guides(colour = "none")

# Combine the plots using patchwork and add a single legend on the right
combined_plot <- ( (plot2 | plot1) / (plot4 | plot3) ) +
  plot_layout(guides = "collect") &
  theme(legend.position = "right")



# Print the combined plot
print(combined_plot)
ggsave("images/fig-sim.svg", width = 13, height = 10, dpi = 300)


save(cox_n0, file = "output/cox_n0.RData")
save(cox_n1, file = "output/cox_n1.RData")
save.image(file = "output/combined_simN.RData")
#rm( list = ls())
```


```{r}  
#| echo: FALSE
#| output: FALSE
#| message: FALSE
#| warning: FALSE
#| label: fig-sim
#| fig-width: 10
#| fig-height: 6.2
#| fig-cap: "Simulation of survival data and effective sample sizes of survival estimates with no censoring. Effective sample sizes for Kaplan-Meier survival estimates are 5,000, the number of patients in both subgroups. There is no decrease because there is no censoring. For the Cox model predictions, the effective sample size begins at 7,500 for men and 15,000 for women. Initially, it decreases with longer follow-up, but later on it increases again."

load("output/combined_simN.RData")
print(combined_plot)
```

```{r}
#| echo: FALSE
#| output: FALSE
#| eval: FALSE
#| message: FALSE
#| warning: FALSE
#| fig-width: 10
#| fig-height: 6.2
#| label: fig-variance
#| fig-cap: "Figure showing the incremental change in the standard error of the baseline cumulative hazards of the Cox model and Kaplan-Meier estimators and showing the incremental change in standard error caused by coefficient misestimation for the Cox model predictions. (Could be better to show cumulative std. err. instead of incremental?)"

load("output/cox_n0.RData")
load("output/cox_n1.RData")
source("code/compile.R")

time <- sort(unique(c(cox_n0$time, cox_n1$time)))
time0 <- sort(unique(c(cox_n0$time)))
time1 <- sort(unique(c(cox_n1$time)))

colors <- c("1" = "#37293F", "3" = "#2E7691",  "2" = "#9f84af", "4" = "#c6aa2c")
linetype <- c("1" = "solid", "2" = "dashed")

plot <- ggplot() +
  geom_line(aes(x = time, y = cox_n0$std.chaz, color = "1", lty = "1"), size = 1, lwd = 1.2) +
  geom_line(aes(x = time, y = cox_n1$std.chaz, color = "2", lty = "1"), size = 1, lwd = 1.2) +
  geom_line(aes(x = time, y = cox_n0$std.coef, color = "1", lty = "2"), size = 1, lwd = 1.2) +
  geom_line(aes(x = time, y = cox_n1$std.coef, color = "2", lty = "2"), size = 1, lwd = 1.2) +
  scale_x_continuous(name = "Time", limits = c(0, 15)) +
  scale_y_continuous(name = "Standard error split by type", limits = c(0, 0.05) ) +
  ggtitle("Standard error of predictions for men and women") +
  labs( color = "Legend") +
  scale_color_manual(
    values = colors,
    labels = c("Cox Men", "Cox Women", "Kaplan-Meier Men", "Kaplan-Meier Women")
  ) +
  scale_linetype_manual(
    values = linetype,
    labels = c( "Cumalative hazard", "Coefficient" ) ) +
  theme_minimal() +
  theme(legend.position = "right") + 
  guides(colour = guide_legend(override.aes = list(alpha = 1)))

plot
```
```{r}
#| echo: FALSE
#| output: FALSE
#| message: FALSE
#| warning: FALSE
#| fig-width: 10
#| fig-height: 6.2
#| label: fig-variance-inc
#| fig-cap: "Figure showing the incremental increases in standard error of the cumulative hazard estimation and coefficient estimation separately. In addtion, it shows the relative contribution to the total variance of Cox model prediction by the estimated baseline. At the start of the follow-up, most of the variance is purely caused by baseline hazard estimation. After a while, the coefficient also contributes substantially. Later in follow-up for the group of men, the baseline hazard starts increasing in contribution again, likely due to a very low remaining survival probability in this group."

load("output/cox_n0.RData")
load("output/cox_n1.RData")
source("code/compile.R")

time <- sort(unique(c(cox_n0$time, cox_n1$time)))
time0 <- sort(unique(c(cox_n0$time)))
time1 <- sort(unique(c(cox_n1$time)))

cox_incremental_hazvar0 <- (c(cox_n0$std.chaz, 0)-c(0, cox_n0$std.chaz))*cox_n0$surv
cox_incremental_hazvar1 <- (c(cox_n1$std.chaz, 0)-c(0, cox_n1$std.chaz))*cox_n0$surv
cox_incremental_coefvar0 <- (c(cox_n0$std.coef, 0)-c(0, cox_n0$std.coef))*cox_n0$surv
cox_incremental_coefvar1 <- (c(cox_n1$std.coef, 0)-c(0, cox_n1$std.coef))*cox_n0$surv
colors <- c("1" = "#37293F", "3" = "#2E7691",  "2" = "#9f84af", "4" = "#c6aa2c")
linetype <- c("1" = "solid", "2" = "dashed", "3" = "dotted")

scaleFactor = 0.00003/1

plot1 <- ggplot() +
  geom_line(aes(x = time, y = cox_incremental_hazvar0[-(length(time)+1)], color = "1", lty = "1"), size = 1, lwd = 1.2) +
  geom_line(aes(x = time, y = cox_incremental_hazvar1[-(length(time)+1)], color = "2", lty = "1"), size = 1, lwd = 1.2) +
  geom_line(aes(x = time, y = cox_incremental_coefvar0[-(length(time)+1)], color = "1", lty = "2"), size = 1, lwd = 1.2) +
  geom_line(aes(x = time, y = cox_incremental_coefvar1[-(length(time)+1)], color = "2", lty = "2"), size = 1, lwd = 1.2) +
  scale_x_continuous(name = "Time", limits = c(0, 15)) +
  scale_y_continuous(name = "Incremental standard error", limits = c(0, 0.000005)) +
  ggtitle("Increment of standard error") +
  labs( color = "Legend") +
  scale_color_manual(
    values = colors,
    labels = c("Women", "Men")
  ) +
  scale_linetype_manual(
    values = linetype,
    labels = c( "Cumalative hazard", "Coefficient", "Reference-Binomial" ) ) +
  theme_minimal() +
  theme(legend.position = "right") + 
  guides(colour = guide_legend(override.aes = list(alpha = 1)))


plot2 <- ggplot() +
  geom_line(aes(x = time, y = cox_n0$std.chaz^2/cox_n0$std.err^2, color = "1", lty = "1"), size = 1, lwd = 1.2) +
  geom_line(aes(x = time, y = cox_n1$std.chaz^2/cox_n1$std.err^2, color = "2", lty = "1"), size = 1, lwd = 1.2) +
  scale_x_continuous(name = "Time", limits = c(0, 15)) +
  scale_y_continuous(name = "Proportion", limits = c(0, 1)) +
  
  ggtitle("Proportion of variance from baseline hazard") +
  labs( color = "Legend") +
  scale_color_manual(
    values = colors,
    labels = c("Women", "Men")
  ) +
  scale_linetype_manual(
    values = linetype,
    labels = c( "Baseline hazard", "Coefficient" ) ) +
  theme_minimal() +
  theme(legend.position = "right") + 
  guides(colour = guide_legend(override.aes = list(alpha = 1)))

# Combine the plots using patchwork and add a single legend on the right
combined_plot <- (plot2 | plot1) +
  plot_layout(guides = "collect") &
  theme(legend.position = "right")

# Print the combined plot
print(combined_plot)
```
```{r}
#| echo: FALSE
#| output: FALSE
#| message: FALSE
#| warning: FALSE
#| fig-width: 10
#| fig-height: 6.2
#| label: fig-variance-contribution
#| fig-cap: "Figure showing the relative contribution to the total variance of Cox model prediction by the estimated baseline and estimated coefficient. At the start of the follow-up, most of the variance is purely caused by baseline hazard estimation. After a while, the coefficient also contributes substantially. Later in follow-up for the group of men, the baseline hazard starts increasing in contribution again, likely due to a very low remaining survival probability in this group."

load("output/cox_n0.RData")
load("output/cox_n1.RData")
library(ggplot2)

time <- sort(unique(c(cox_n0$time, cox_n1$time)))
time0 <- sort(unique(c(cox_n0$time)))
time1 <- sort(unique(c(cox_n1$time)))

colors <- c("1" = "#37293F", "3" = "#2E7691",  "2" = "#9f84af", "4" = "#c6aa2c")
linetype <- c("1" = "solid", "2" = "dashed")

plot <- ggplot() +
  geom_line(aes(x = time, y = cox_n0$std.chaz^2/cox_n0$std.err^2, color = "1", lty = "1"), size = 1, lwd = 1.2) +
  geom_line(aes(x = time, y = cox_n1$std.chaz^2/cox_n1$std.err^2, color = "2", lty = "1"), size = 1, lwd = 1.2) +
  geom_line(aes(x = time, y = cox_n0$std.coef^2/cox_n0$std.err^2, color = "1", lty = "2"), size = 1, lwd = 1.2) +
  geom_line(aes(x = time, y = cox_n1$std.coef^2/cox_n1$std.err^2, color = "2", lty = "2"), size = 1, lwd = 1.2) +
  scale_x_continuous(name = "Time", limits = c(0, 1.5)) +
  scale_y_continuous(name = "Proportion of total variance", limits = c(0, 1)) +
  ggtitle("Contribution of hazard and coefficient to total variance") +
  labs( color = "Legend") +
  scale_color_manual(
    values = colors,
    labels = c("Women", "Men")
  ) +
  scale_linetype_manual(
    values = linetype,
    labels = c( "Baseline hazard", "Coefficient" ) ) +
  theme_minimal() +
  theme(legend.position = "right") + 
  guides(colour = guide_legend(override.aes = list(alpha = 1)))

plot
```


```{r}  
#| echo: FALSE
#| output: FALSE
#| eval: FALSE
#| message: FALSE
#| warning: FALSE
#| fig-width: 10
#| fig-height: 6.2
#| label: fig-sim-dev-cont
#| fig-cap: "Simulation of survival data and effective sample sizes of survival estimates with no censoring and one continuous predictor."


source("code/compile.R")
require(surveillance)

# Seed for consistency
set.seed(1923)

scale = c(10)
shape = c(1) # shape 1 means exponential
maxtime = 100 # maxtime
hr = 2 # hazard ratio
n = 10000 # number per group

x <- data.frame(x = rnorm(n, 0, 1))
events <- simsurv( dist="weibull", 
gammas=shape, 
lambdas=scale[1]^(-shape), 
x=x,
betas = c("x" = log(hr)),
maxt=maxtime)
events$x = x$x
time <- sort(unique(events$eventtime))

# Fit the models
cox <- coxph(Surv(eventtime, status) ~ x, data = events)

# Calculate effective N and stuff
cox_mean <- survfit_n(survfit(cox, newdata = data.frame(x = 0)), cox, coef=T, chaz=T)
cox_sd_plus <- survfit_n(survfit(cox, newdata = data.frame(x = 1)), cox, coef=T, chaz=T)
cox_sd_min <- survfit_n(survfit(cox, newdata = data.frame(x = -1)), cox, coef=T, chaz=T)

cox_mean <- sf_to_df(cox_mean, time = time)
cox_sd_plus <- sf_to_df(cox_sd_plus, time = time)
cox_sd_min <- sf_to_df(cox_sd_min, time = time)

colors <- c("1" = "#37293F", "3" = "#2E7691",  "2" ="#C2666B", "4" = "#c6aa2c")
linetype <- c("1" = "solid", "2" = "dashed")

# plot the effective sample sizes all in one plot
plot1 <- ggplot() +
geom_line(aes(x = time, y = cox_mean$n.eff, color = "1"), size = 1.2) +
geom_line(aes(x = time, y = cox_sd_plus$n.eff, color = "2"), size = 1.2) +
geom_line(aes(x = time, y = cox_sd_min$n.eff, color = "3"), size = 1.2) +
scale_x_continuous(name = NULL, labels = NULL, limits = c(0, 15)) +
scale_y_continuous(name = "Effective Sample Size (N)", limits = c(0, 15000)) +
#ggtitle("Effective Sample Size of Coxph and Survfit") +
labs( color = "Legend", tag = "(B)") +
scale_color_manual(
values = colors,
labels = c("Mean", "+1 SD", "-1 SD")
) +
theme_minimal() +
theme(legend.position = "right") + 
guides(colour = guide_legend(override.aes = list(alpha = 1)))


# plot the two kaplan-meiers overlapping in one plot and the cox survival predictions for the 0 an 1 treatment patients
plot2 <- ggplot() +
geom_line(aes(x = time, y = cox_mean$surv*100, color = "1"), size = 1.2) +
geom_line(aes(x = time, y = cox_sd_plus$surv*100, color = "2"), size = 1.2) +
geom_line(aes(x = time, y = cox_sd_min$surv*100, color = "3"), size = 1.2) +
scale_x_continuous(name = "Time", limits = c(0, 15)) +
scale_y_continuous(name = "Predicted survival (%)", limits = c(0, 100)) +
#ggtitle("Survival of Coxph and Survfit") +
labs( color = "Legend", tag = "(A)") +
scale_color_manual(
values = colors,
labels = c("Mean", "+1 SD", "-1 SD")
) +
theme_minimal() +
theme(legend.position = "none") + 
guides(colour = guide_legend(override.aes = list(alpha = 1))) 

scaleFactor = 0.00003/1

plot3 <- ggplot() +
geom_line(aes(x = time, y = cox_mean$std.chaz^2, color = "1", lty = "1"), size = 1, lwd = 1.2) +
geom_line(aes(x = time, y = cox_sd_plus$std.chaz^2, color = "2", lty = "1"), size = 1, lwd = 1.2) +
geom_line(aes(x = time, y = cox_sd_min$std.chaz^2, color = "3", lty = "1"), size = 1, lwd = 1.2) +
geom_line(aes(x = time, y = cox_mean$std.err^2, color = "1", lty = "2"), size = 1, lwd = 1.2) +
geom_line(aes(x = time, y = cox_sd_plus$std.err^2, color = "2", lty = "2"), size = 1, lwd = 1.2) +
geom_line(aes(x = time, y = cox_sd_min$std.err^2, color = "3", lty = "2"), size = 1, lwd = 1.2) +
scale_x_continuous(name = "Time", limits = c(0, 15)) +
scale_y_continuous(name = "Variance of Cox model", limits = c(0, 0.003)) +
#ggtitle("Increment of standard error") +
labs( color = "Legend", tag = "(D)") +
scale_color_manual(
values = colors,
labels = c("Mean", "+1 SD", "-1 SD")
) +
scale_linetype_manual(
values = linetype,
labels = c( "Baseline hazard", "Total variance" ) ) +
theme_minimal() +
theme(legend.position = "none") + 
guides(colour = guide_legend(override.aes = list(alpha = 1)))


plot4 <- ggplot() +
geom_line(aes(x = time, y = cox_mean$std.chaz^2/cox_mean$std.err^2*100, color = "1"), size = 1, lwd = 1.2) +
geom_line(aes(x = time, y = cox_sd_plus$std.chaz^2/cox_sd_plus$std.err^2*100, color = "2"), size = 1, lwd = 1.2) +
geom_line(aes(x = time, y = cox_sd_min$std.chaz^2/cox_sd_min$std.err^2*100, color = "3"), size = 1, lwd = 1.2) +
scale_x_continuous(name = "Time", limits = c(0, 15)) +
scale_y_continuous(name = "Percentage of variance from h0", limits = c(0, 100)) +

#ggtitle("Proportion of variance from baseline hazard") +
labs( color = "Legend", tag = "(C)") +
scale_color_manual(
values = colors,
labels = c("Mean", "+1 SD", "-1 SD")
) +
theme_minimal() +  
theme(legend.position = "right") + 
guides(colour = "none")

# Combine the plots using patchwork and add a single legend on the right
combined_plot <- ( (plot2 | plot1) / (plot4 | plot3) ) +
plot_layout(guides = "collect") &
theme(legend.position = "right")



# Print the combined plot
print(combined_plot)
ggsave("images/fig-sim-cont.svg", width = 13, height = 10, dpi = 300)


#save(cox_n0, file = "output/cox_n0.RData")
#save(cox_n1, file = "output/cox_n1.RData")
save.image(file = "output/combined_simN_cont.RData")
#rm( list = ls())
```

#### Heuristic explanation: Baseline hazard estimation
The baseline hazard for the Cox model at time $t_i$ with $\delta_i$ events is $\frac{\delta_i}{\sum_{j\in R_i}\exp(\beta'X_j)}$, where the denominator is the risk-weighted number at risk. This risk-weighted number at risk is the implicit sample size used for the baseline hazard estimation. A heuristic explanation for the effective sample sizes observed near $t=0$ in @fig-sim would be the following. Assuming the hazard ratio was correctly estimated by the model and there has been a single death in 10,000 patients. The risk-weighted number at risk in the data is $5,000\cdot2 + 5,000\cdot 1 = 15,000$. If you are a woman, your risk of being this first death are $\frac{1}{15,000}$, giving us an effective sample size of 15,000. If you are a man, your risk of being this first death are $\frac{2}{15,000} = \frac{1}{7,500}$, giving us an effective sample size of 7,500. 

We can check this explanation in our simulated data. On the l.h.s. of @fig-variance-inc, we can see that indeed most of the contribution to the variance early on in follow-up is from the baseline hazard estimation. Additionally, with the Cox-model estimated HR of 1.98, the expected effective sample sizes from our explanation would be 14910 for women and 7523 for men. The estimated effective sample sizes are 14907 and 7523 respectively, where the small difference for women can be explained by the contribution of the coefficient estimation. Details can be read in Supplement 6.

#### Heuristic explanation: Coefficient contribution
In @fig-variance-inc, we can see that the contribution of the coefficient estimation to the total variance is somewhat similar between the two groups in relation to the baseline hazard contribution. At $t=7.5$ we can clearly see a divergence of this pattern, where the contribution of the coefficient tot the total variance for women drops off substantially. Using the formula from @eq-cox_var_wp we can explain this.

At the time of the first event, the gradient (@eq-cox_var_grad) for both patients is $\frac{2}{45000}$ and the variance of the coefficient is constant. Therefore the uncertainty from possibly wrongly estimating the coefficient is the same for both groups. Calculating the absolute number at later time points is more difficult, because the gradient is a cumulative sum over all event times. We can look at the newest term of this sum in the gradient at time $t_i$. We choose $t=6.931$, which is the exact time where $S_{women}(6.931)=0.5$ and $S_{men}(6.931)=0.25$ (Supplement 7). If an event occurs at this time, the expected new addition to the cumulative sum of the gradient is $\frac{1}{10000}$ for women and $-\frac{2}{10000}$ for men because of the drift in the number of patients at risk from each of the groups. 


<!-- ## Parametric model -->
<!-- A final simulation step is to compare the results of the Cox model effective sample size to that of proportional hazards parametric survival model, both well specified and misspecified. Our simulated data was from a Weibull(2, 1) hazard distribution with a hazard ratio of 2. First we fit a Weibull proportional hazards model, which is correctly specified. -->
```{r}
#| echo: FALSE
#| output: FALSE
#| eval: FALSE
#| message: FALSE
#| warning: FALSE
#| fig-width: 10
#| fig-height: 6.2
#| label: fig-sim-well-specified
#| fig-cap: "Simulation of survival data and effective sample sizes of survival estimates with no censoring. Effective sample size for Kaplan-Meier survival estimates is 5,000 all the time for both groups. For the Cox model predictions, the effective sample size begins at 7500 for men and 15000 for women. The explaination for this is that the weighted risk set is 7500 in the Cox model. When the first patient dies, a man has a 1/7500 chance of mortality, showing the effective sample size of 7500. A woman has a 0.5/7500 = 15000 risk of death, showing an assumed effective sample size of 15000."


source("code/compile.R")
require(surveillance)
require(flexsurv)

# Seed for consistency
set.seed(1923)

scale = c(1)
shape = c(1) # shape 1 means exponential
maxtime = 10 # maxtime
hr = 2 # hazard ratio
n = 5000 # number per group

events <- simsurv( dist="weibull", 
                   gammas=shape, 
                   lambdas=scale[1]^(-shape), 
                   x=data.frame(tx = c(rep(0, n), rep(1, n))),
                   betas = c("tx" = log(hr)),
                   maxt=maxtime)
events$tx = c(rep(0, n), rep(1, n))
events$eventtime <- untie(events$eventtime)
time <- sort(unique(events$eventtime))
time0 <- sort(unique(events[events$tx == 0, ]$eventtime))
time1 <- sort(unique(events[events$tx == 1, ]$eventtime))

# Fit the models
sf0 <- survfit(Surv(eventtime, status) ~ 1, data = events[events$tx == 0,])
sf1 <- survfit(Surv(eventtime, status) ~ 1, data = events[events$tx == 1,])
weibull <- flexsurvreg(Surv(eventtime, status) ~ tx, data = events, dist = "exp")
pred <- summary(weibull, newdata = data.frame(tx = c(0, 1)), type = "survival", t = time, se=TRUE)

# Calculate effective N and stuff
sf_n0 <- survfit_n(sf0) |> sf_to_df(time = time0)
sf_n1 <- survfit_n(sf1) |> sf_to_df(time = time1)
weibull_n0 <- pred[[1]]$est*(1-pred[[1]]$est)/(pred[[1]]$se^2)
weibull_n1 <- pred[[2]]$est*(1-pred[[2]]$est)/(pred[[2]]$se^2)

colors <- c("1" = "#37293F", "3" = "#2E7691",  "2" = "#9f84af", "4" = "#c6aa2c")
linetype <- c("1" = "solid", "2" = "dashed")

# plot the effective sample sizes all in one plot
plot1 <- ggplot() +
  geom_line(aes(x = time, y = weibull_n0, color = "1"), size = 1.2) +
  geom_line(aes(x = time, y = weibull_n1, color = "2"), size = 1.2) +
  geom_line(aes(x = time1, y = sf_n0$n.eff, color = "3"), size = 1.2) +
  geom_line(aes(x = time1, y = sf_n1$n.eff, color = "4"), size = 1.2) +
  scale_x_continuous(name = "Time", limits = c(0, 1.5)) +
  scale_y_continuous(name = "Effective Sample Size", limits = c(0, 50000)) +
  ggtitle("Effective Sample Size of Exponential PH and Survfit") +
  labs( color = "Legend") +
  scale_color_manual(
    values = colors,
    labels = c("Exp-PH Women", "Exp-PH Men", "KM Women", "KM Men")
  ) +
  theme_minimal() +
  theme(legend.position = "right") + 
  guides(colour = guide_legend(override.aes = list(alpha = 1)))


# plot the two kaplan-meiers overlapping in one plot and the cox survival predictions for the 0 an 1 treatment patients
plot2 <- ggplot() +
  geom_line(aes(x = time0, y = sf_n0$surv, color = "3"), size = 1.2) +
  geom_line(aes(x = time1, y = sf_n1$surv, color = "4"), size = 1.2) +
  geom_line(aes(x = time, y = pred[[1]]$est, color = "1"), size = 1.2) +
  geom_line(aes(x = time, y = pred[[1]]$est, color = "2"), size = 1.2) +
  scale_x_continuous(name = "Time", limits = c(0, 1.5)) +
  scale_y_continuous(name = "Survival") +
  ggtitle("Survival of Exponential PH and Survfit") +
  labs( color = "Legend") +
  scale_color_manual(
    values = colors,
    labels = c("Exp-PH Women", "Exp-PH Men", "KM Women", "KM Men")
  ) +
  theme_minimal() +
  theme(legend.position = "right") + 
  guides(colour = guide_legend(override.aes = list(alpha = 1)))


# Combine the plots using patchwork and add a single legend on the right
combined_plot_exp <- (plot2 | plot1) +
  plot_layout(guides = "collect") &
  theme(legend.position = "right")

# Print the combined plot
print(combined_plot_exp)

#exp(cox$coefficients)
print(weibull_n0[100])
print(weibull_n1[100])

#save(cox_n0, file = "output/cox_n0.RData")
#save(cox_n1, file = "output/cox_n1.RData")
save.image(file = "output/combined_exp.RData")
#rm( list = ls())
```

<!-- ### Confidence bands -->
<!-- Voorstel van Liesbeth - leek niet te lukken (stond dat er geen goede software is in @sachs2022), maar riskRegression heeft wel een implementatie. Wel vreemd, effective N op basis van hun SE is heel anders dat die van `survival`. riskRegression zegt dat ze een von Mises expansion gebruiken voor de SE (in plaats van de delta-methode). Dit ook nog toevoegen lijkt me wel erg veel werk en niet een gigantisch verschil. Deze SE zorgt voor exact dezelfde effective N super vroeg, daarna ietsje lager en later weer gelijk. Ander apart iets - de effective N op basis van deze SE gaat heel vaak ook omhoog - ik denk wanneer iemand gecensureerd wordt ipv overlijd of andersom? Niet helemaal duidelijk nog, maar zelfs na een tijdje gebeurt dit nog. Er worden ook geen "SE" voor de confidence bands gerapporteerd, dus het beste dat je kan doen is de breedte delen door 2*1.96 en dat gebruiken als SE. Dit lijkt best goed te gaan bij de pointwise CI. -->

<!-- ```{r} -->
<!-- source("code/compile.R") -->
<!-- library(riskRegression) -->
<!-- cox <- coxph(Surv(time, status)~ age, data = colon_df, x = TRUE) -->
<!-- pred <- predictCox(cox, time = sort(colon_df$time), newdata = data.frame(age = 70), band = TRUE, se = TRUE) -->
<!-- plot <- ggplot() + -->
<!--   geom_line(aes(x = pred$times, y = pred$survival[1,]), linewidth = 1.2) + -->
<!--   geom_line(aes(x = pred$times, y = pred$survival.upper[1,]), linetype = "dashed") + -->
<!--   geom_line(aes(x = pred$times, y = pred$survival.lower[1,]), linetype = "dashed") + -->
<!--   geom_line(aes(x = pred$times, y = pred$survival.upperBand[1,]), linetype = "dotted") + -->
<!--   geom_line(aes(x = pred$times, y = pred$survival.lowerBand[1,]), linetype = "dotted") + -->
<!--   scale_x_continuous(name = "Time") + -->
<!--   scale_y_continuous(name = "Survival Probability", limits = c(0, 1)) + -->
<!--   ggtitle("Cox model prediction with pointwise confidence interval") + -->
<!--   theme_minimal() -->
<!-- plot -->

<!-- n_band <- pred$survival[1,]*(1-pred$survival[1,])/( ((pred$survival.upperBand[1,] - pred$survival.lowerBand[1,])/(2*1.96))^2 ) -->
<!-- n_int <- pred$survival[1,]*(1-pred$survival[1,])/( ((pred$survival.upper[1,] - pred$survival.lower[1,])/(2*1.96))^2 ) -->
<!-- n_se <- pred$survival[1,]*(1-pred$survival[1,])/( (pred$survival.se[1,])^2 ) -->

<!-- sf_n <-  survfit_n(survfit(cox, newdata = data.frame(age = 70)), cox = cox) -->

<!-- plot2 <- ggplot() + -->
<!--   geom_line(aes(x = pred$times, y = n_se, color = "1"), linewidth = 1.2) + -->
<!--   geom_line(aes(x = pred$times, y = n_int, color = "2"), linewidth = 1.2) + -->
<!--   geom_line(aes(x = pred$times, y = n_band, color = "3"), linewidth = 1.2) + -->
<!--   geom_line(aes(x = sf_n$time, y = sf_n$n.eff, color = "4"), linewidth = 1.2) + -->
<!--   geom_hline(yintercept = 929, linetype = "dotted", color = "black", linewidth = 0.6) + -->
<!--   scale_x_continuous(name = "Time") + -->
<!--   scale_y_continuous(name = "Effective Sample Size", limits = c(0, 1000)) + -->
<!--   ggtitle("Effective Sample Size from riskRegression") + -->
<!--   labs( color = "Legend") + -->
<!--   scale_color_manual( -->
<!--     values = c("1"="#37293F", "2"="#2E7691", "3"="#9f84af", "4"="#c6aa2c"), -->
<!--     labels = c("Using standard error", "Using pointwise CI", "Using simultaneous CI", "Using survfit_n") -->
<!--   ) + -->
<!--   theme_minimal() + -->
<!--   theme(legend.position = "right") +  -->
<!--   guides(colour = guide_legend(override.aes = list(alpha = 1))) -->
<!-- plot2 -->
<!-- ``` -->

## Discussion
This study shows that effective sample size behaves in a complicated way for the Cox proportional hazards model. We relate our findings to the criteria set in the introduction.
1.	Natural behavior: Effective sample size for Cox model predictions only partly reflects the size of a sample with complete follow-up, accommodating censoring. The effective sample size starts out higher and decreases with longer follow-up. Without censoring, it decreases towards the effective sample size of a Kaplan-Meier estimate in a Cox model with one categorical variable. With censoring, effective sample size also decreases in line with the pattern seen in Kaplan-Meier effective sample size. Additionally, effective sample size can also increase with longer follow-up for patients with high hazard ratios.
2.	Reflection of exceptionality: Effective sample size reflects the commonality of a patient profile within the development data. This criterium is met as shown in Figure 3. Patients with common profiles, such as the median age, have an effective sample size equal to the development sample size, while patients with exceptional age (99th percentile) have a low effective sample size.


#### Normality and linearity assumptions - ignore for now
Effective sample size in linear regression is bounded by the total sample size.[@thomassen2024] For prediction from a linear regression model with multiple variables, a patient with exactly the population mean covariate values will have exactly the total development sample as effective sample size (Supplement 2). We might anticipate / expect that we are as certain of the event probability for such an average patient. 



#### Issues with variance estimation when the event probability is very high or low
We may assume that the outcome variance behaves according to a binomial distribution. with variance p(1-p). This variance becomes very small with very high or very low event probabilities, inflating effective sample sizes. This phenomenon was observed for very low risk predictions based on a logistic regression model[@thomassen2024] and Cox models for high and low event probabilities (Figure 3 and Figure 4).

### Communication of effective sample size
Effective sample size may be valuable for the communication of uncertainty in clinical predictions with patients. It could be an addition or a replacement of a 95% confidence interval. Especially when a risk is communicated as “if we had 100 patients like you, we would expect X to survive”,[@kattan2002] effective sample size could be used to specify that it is not based on 100, but more or fewer patients.

For communication with patients, the behavior early in follow-up of effective sample size for the Cox model predictions may be problematic. It is counterintuitive and could be considered misleading. While the behavior can be explained (section 4.2.1), this does not prevent misinterpretation.

However, the behavior early in follow-up may be less relevant than for predictions at later time points. In clinical fields such as cardiovascular disease, prediction models commonly have 10-year time horizons.[@visseren2021] Effective sample size behaves in line with our ‘natural behavior’ criterium for later follow-up times (Figure 3). 


### Cox model variance is too low
When the effective sample size could be considered unreasonably large, the variance and resulting 95% pointwise confidence interval of the prediction could also be considered unreasonably small. Limited studies exist that investigate the confidence interval coverage of Cox model predictions, especially at the early times where effective sample size is higher than the development sample. It is known that Wald intervals for the survival function are rather poor[@burr1994] and they can be much improved through log or log-log transformations.[@link1984;@andersen1996] It is peculiar that the variance of a Cox model prediction for a patient with rather unique covariate values can be much larger than the total sample size early in follow-up. This implies that many more patients would be required for a similar level of certainty if a logistic regression model was used instead. It appears that much of this is caused by the low uncertainty of the baseline hazard early on in follow-up. The success of the Cox model may have had the unintended consequence of reduced effort towards studying the baseline hazard.[@hjort1992] Flexible parametric models have been suggested as an alternative that allows for a less erratic and more informative baseline hazard.[@royston2002] Studies comparing Cox models to parametric survival models have mostly focused on correct estimation of the coefficients.[@nardi2003] For the purposes of a survival prediction model, the absolute risk often matters most, with the baseline hazard as an essential component.

```{r}
#| echo: FALSE
#| output: FALSE
#| eval: FALSE
#| message: FALSE
#| warning: FALSE
#| fig-width: 10
#| fig-height: 6.2
#| label: fig-sim-coverage
#| fig-cap: "Simulation of the coverage of a Cox model prediction."


source("code/compile.R")
require(surveillance)

# Seed for consistency
set.seed(1923)

scale = c(10)
shape = c(1) # shape 1 means exponential
maxtime = 100 # maxtime
hr = 2 # hazard ratio
n = 5000 # number per group

events <- simsurv( dist="weibull", 
gammas=shape, 
lambdas=scale[1]^(-shape), 
x=data.frame(tx = c(rep(0, n), rep(1, n))),
betas = c("tx" = log(hr)),
maxt=maxtime)
events$tx = c(rep(0, n), rep(1, n))
events$eventtime[8885] <- 2.997671  # handmatig ties weghalen
events$eventtime[6298] <- 5.668818  # werkt omdat seed vast staats
time <- sort(unique(events$eventtime))
time0 <- sort(unique(events[events$tx == 0, ]$eventtime))
time1 <- sort(unique(events[events$tx == 1, ]$eventtime))

# Fit the models
sf0 <- survfit(Surv(eventtime, status) ~ 1, data = events[events$tx == 0,])
sf1 <- survfit(Surv(eventtime, status) ~ 1, data = events[events$tx == 1,])
cox <- coxph(Surv(eventtime, status) ~ tx, data = events)

# Calculate effective N and stuff
sf_n0 <- survfit_n(sf0) |> sf_to_df(time = time0)
sf_n1 <- survfit_n(sf1) |> sf_to_df(time = time1)
cox_n0 <- survfit_n(survfit(cox, newdata = data.frame(tx = 0)), cox, coef=T, chaz=T)
cox_n1 <- survfit_n(survfit(cox, newdata = data.frame(tx = 1)), cox, coef=T, chaz=T)

cox_n0 <- sf_to_df(cox_n0, time = time)
cox_n1 <- sf_to_df(cox_n1, time = time)

colors <- c("1" = "#37293F", "3" = "#2E7691",  "2" = "#9f84af", "4" = "#c6aa2c")
linetype <- c("1" = "solid", "2" = "dashed")

coverage = data.frame(time = c(0.01, 0.1, 0.5, 1, 5, 10), Neff_men = cox_n0[which.min(time<)]), coverage_men = NA, coverage_women = NA)
for( B in 1:1000 ){
outcomes_men <- simsurv( dist = "weibull",
gammas = shape, 
lambdas = scale[1]^(-shape), 
x = data.frame(tx = rep(1, cox_n1$n.eff[1])),
betas = c("tx" = log(hr)),
maxt = maxtime)
outcomes_women <- simsurv( dist = "weibull",
gammas = shape, 
lambdas = scale[1]^(-shape), 
x = data.frame(tx = rep(0, cox_n0$n.eff[1])),
betas = c("tx" = log(hr)),
maxt = maxtime)

}

```

### Effective sample size in model development
In Figure 3, we note that the addition of covariates to a Cox model has differential impacts on the spread of predictions and effective sample sizes. This implies that a balance must be struck between bias and variance to make predictions for individual patients. The addition of age to the treatment-only model, for example, only changed the predictions slightly, but decreased the effective sample size for older and younger patients substantially. This trade-off must be considered when developing prediction models for informing individual patients. Effective sample size provides insight into such trade-offs. Effective sample size may be valuable in addition to other measures to indicate model performance such as the improvement in model fit, as expressed with $R^2$ values or Brier score.[@vancalster2025]

Additionally, certain patients may have extremely low effective sample sizes in the full model (Figure 3). Such low effective sample sizes imply that their predictions are too uncertain to be relied upon. Modelers may consider winsorizing the covariate values to obtain more certain predictions.[@royston2000;@aguinis2013] Moreover, variants of penalized prediction could be developed to stabilize predictions especially in the tails of continuous predictor values.[@houwelingen1990;@verweij1994] For example, elastic net regression has much higher minimum effective sample size than logistic regression.[@thomassen2025]

Guidelines for the distribution of effective sample size are difficult to decide on, as much of it is context dependent. Suggestions could be a median effective sample size of at least 100 to ensure enough certainty for the majority of patients[@harrellblog] and a minimum effective sample size of at least 10.[@thomassen2024]


## Conclusion
Effective sample size for the Cox model is an complicated measure for the uncertainty of individual survival predictions. Early on in follow-up, the survival probability and uncertainty are dominated by the estimated baseline hazard, such that the effective sample size for specific patient profiles can be higher than the total sample size. This reflects that the Cox model prediction variance is lower than the outcome variance in a sample of $N_{eff}$ patients with the same covariate values with complete follow-up.

Hence, effective sample size can be misleading for short follow-up times. Effective sample size for Cox model predictions is more meaningful for late follow-up times. Further research is needed into the understanding of measures for uncertainty before communication with effective sample size is implemented in clinical practice.


# New simulation study

```{r}
source("code/compile.R")
require(surveillance)

# Seed for consistency
set.seed(1923)

shape = c(0.5) # shape 1 means exponential
scale = c(lambda(0.25, shape), lambda(0.5, shape), lambda(0.75, shape), lambda(0.9, shape), lambda(0.95, shape)) # scale
maxtime = 10 # maxtime
hr = 2 # hazard ratio
n = 1000 # number per group
time_grid <- seq(0, 10, by = 0.001)
n_sim <- 100  # choose number of repeats

df_standard <- bind_rows(
  lapply(1:n_sim, function(r)
    run_one(r, scale, shape, maxtime, hr, n, time_grid, model_type = "standard")
  )
)

df_standard$prevalence <- factor(df_standard$prevalence, levels = c("75%", "50%", "25%", "10%", "5%"))
df_standard <- df_standard %>%
  group_by(tx, prevalence) %>%
  mutate(events = max(n.risk) - n.risk)
df_summary_std <- df_standard %>%
  group_by(time, tx, prevalence) %>%
  summarise(
    n.eff = mean(n.eff),
    surv = mean(surv),
    events = mean(events),
    .groups = "drop"
  )  %>%
  ungroup() %>%
  group_by(tx, prevalence) %>%
  mutate(events_rate = events / max(events),
         surv_prop = (1-surv)/max(1-surv)) %>%
  ungroup()





colors <- c("0" = "#37293F", "1" = "#9f84af")

linetypes <- c(
  "75%" = "twodash",
  "50%" = "solid",
  "25%" = "dashed",
  "10%" = "dotted",
  "5%" = "dotdash"
)

p_time <- ggplot(df_summary_std, aes(x = time, y = n.eff,
                         color = tx,
                         linetype = prevalence,
                         group = interaction(tx, prevalence))) +
  geom_hline(yintercept = 1000, linetype = "dotted", color = "grey50") +
  geom_hline(yintercept = 500,  linetype = "dotted", color = "grey50") +
  geom_line(linewidth = 1) +
  scale_color_manual(values = colors, name = "Group") +
  scale_linetype_manual(values = linetypes, name = "Baseline event rate") +
  labs(x = "Time", y = "Effective sample size") +
  scale_x_continuous(limits = c(0, 10)) +
  scale_y_continuous(limits = c(0, 1600)) +
  theme_minimal()

p_event <- ggplot(df_summary_std, aes(x = events, y = n.eff,
                           color = tx,
                           linetype = prevalence,
                           group = interaction(tx, prevalence))) +
  geom_line(linewidth = 1) +
  geom_hline(yintercept = 1000, linetype = "dotted", color = "grey50") +
  geom_hline(yintercept = 500,  linetype = "dotted", color = "grey50") +
  scale_color_manual(values = colors, name = "Group") +
  scale_linetype_manual(values = linetypes, name = "Baseline event rate") +
  labs(x = "Number of events") +
  scale_x_continuous(limits = c(0, 1000)) +
  scale_y_continuous(limits = c(0, 1600)) +
  theme_minimal() +
  theme( axis.text.y = element_blank(), 
         axis.ticks.y = element_blank(),
         axis.title.y = element_blank())


p_event_prop <- ggplot(df_summary_std, aes(x = events_rate, y = n.eff,
                           color = tx,
                           linetype = prevalence,
                           group = interaction(tx, prevalence))) +
  geom_line(linewidth = 1) +
  geom_hline(yintercept = 1000, linetype = "dotted", color = "grey50") +
  geom_hline(yintercept = 500,  linetype = "dotted", color = "grey50") +
  scale_color_manual(values = colors, name = "Group") +
  scale_linetype_manual(values = linetypes, name = "Baseline event rate") +
  labs(x = "Proportion of total events") +
  scale_x_continuous(limits = c(0, 1)) +
  scale_y_continuous(limits = c(0, 1600)) +
  theme_minimal() +
  theme( axis.text.y = element_blank(), 
         axis.ticks.y = element_blank(),
         axis.title.y = element_blank()) 



df_large <- bind_rows(
  lapply(1:100, function(r)
    run_one(r, scale, shape, maxtime, hr, 10000, time_grid, model_type = "standard")
  )
)

df_large$prevalence <- factor(df_large$prevalence, levels = c("75%", "50%", "25%", "10%", "5%"))
df_large <- df_large %>%
  group_by(tx, prevalence) %>%
  mutate(events = max(n.risk) - n.risk)
df_summary_large <- df_large %>%
  group_by(time, tx, prevalence) %>%
  summarise(
    n.eff = mean(n.eff),
    surv = mean(surv),
    events = mean(events),
    .groups = "drop"
  )  %>%
  ungroup() %>%
  group_by(tx, prevalence) %>%
  mutate(events_rate = events / max(events),
         surv_prop = (1-surv)/max(1-surv)) %>%
  ungroup()


p_time_large <- ggplot(df_summary_large, aes(x = time, y = n.eff,
                         color = tx,
                         linetype = prevalence,
                         group = interaction(tx, prevalence))) +
  geom_line(linewidth = 1) +
  geom_hline(yintercept = 10000, linetype = "dotted", color = "grey50") +
  geom_hline(yintercept = 5000,  linetype = "dotted", color = "grey50") +
  scale_color_manual(values = colors, name = "Group") +
  scale_linetype_manual(values = linetypes, name = "Baseline event rate") +
  labs(x = "Time", y = "Effective sample size") +
    scale_x_continuous(limits = c(0, 10)) +
  scale_y_continuous(limits = c(0, 16000)) +
  theme_minimal()

p_event_large <- ggplot(df_summary_large, aes(x = events, y = n.eff,
                           color = tx,
                           linetype = prevalence,
                           group = interaction(tx, prevalence))) +
  geom_line(linewidth = 1) +
  geom_hline(yintercept = 10000, linetype = "dotted", color = "grey50") +
  geom_hline(yintercept = 5000,  linetype = "dotted", color = "grey50") +
  scale_color_manual(values = colors, name = "Group") +
  scale_linetype_manual(values = linetypes, name = "Baseline event rate") +
  labs(x = "Number of events") +
    scale_x_continuous(limits = c(0, 10000)) +
  scale_y_continuous(limits = c(0, 16000)) +
  theme_minimal() +
  theme( axis.text.y = element_blank(), 
         axis.ticks.y = element_blank(),
         axis.title.y = element_blank())


p_event_prop_large <- ggplot(df_summary_large, aes(x = events_rate, y = n.eff,
                         color = tx,
                         linetype = prevalence,
                         group = interaction(tx, prevalence))) +
  geom_line(linewidth = 1) +
  geom_hline(yintercept = 10000, linetype = "dotted", color = "grey50") +
  geom_hline(yintercept = 5000,  linetype = "dotted", color = "grey50") +
  scale_color_manual(values = colors, name = "Group") +
  scale_linetype_manual(values = linetypes, name = "Baseline event rate") +
  labs(x = "Proportion of final events") +
  scale_x_continuous(limits = c(0, 1)) +
  scale_y_continuous(limits = c(0, 16000)) +
  theme_minimal() +
  theme( axis.text.y = element_blank(), 
         axis.ticks.y = element_blank(),
         axis.title.y = element_blank())

p_surv_large <- ggplot(df_summary_large, aes(x = 1-surv, y = n.eff,
                         color = tx,
                         linetype = prevalence,
                         group = interaction(tx, prevalence))) +
  geom_line(linewidth = 1) +
  scale_color_manual(values = colors, name = "Group") +
  scale_linetype_manual(values = linetypes, name = "Baseline event rate") +
  labs(x = "Event incidence", y = "Effective sample size") +
    scale_x_continuous(limits = c(0, 1)) +
  scale_y_continuous(limits = c(0, 16000)) +
  theme_minimal()


# Combine the plots using patchwork and add a single legend on the right
combined_plot <- ( ( p_time | p_event | p_event_prop )  ) +
  plot_layout(guides = "collect") +
  plot_annotation(
    title = "Effective sample size for different event rates",
    theme = theme(plot.title = element_text(size = 14, face = "bold", hjust = 0))
  ) +
  theme(legend.position = "right")

# Combine the plots using patchwork and add a single legend on the right
combined_plot_large <- ( ( p_time_large | p_event_large | p_event_prop_large )  ) +
  plot_layout(guides = "collect") +
  plot_annotation(
    title = "Effective sample size for different event rates with larger development sample size (N=10,000)",
    theme = theme(plot.title = element_text(size = 14, face = "bold", hjust = 0))
  ) +
  theme(legend.position = "right")

# Print the combined plot
print(combined_plot)
print(combined_plot_large)
```

```{r} 
# Seed for consistency
set.seed(1923)

shape = c(0.5) # shape 1 means exponential
scale = c(lambda(0.25, shape), lambda(0.5, shape), lambda(0.75, shape), lambda(0.9, shape), lambda(0.95, shape)) # scale
maxtime = 10 # maxtime
hr = 2 # hazard ratio
n = 1000 # number per group
time_grid <- seq(0, 10, by = 0.001)
n_sim <- 10  # choose number of repeats

df_robust <- bind_rows(
  lapply(1:n_sim, function(r)
    run_one(r, scale, shape, maxtime, hr, n, time_grid, model_type = "robust")
  )
)

df_robust$prevalence <- factor(df_robust$prevalence,
                               levels = c(0.25, 0.5, 0.75, 0.9, 0.95), 
                               labels = c("75%", "50%", "25%", "10%", "5%"))
df_robust <- df_robust %>%
  group_by(tx, prevalence) %>%
  mutate(events = max(n.risk) - n.risk)
df_summary_rob <- df_robust %>%
  group_by(time, tx, prevalence) %>%
  summarise(
    n.eff = mean(n.eff),
    surv = mean(surv),
    events = mean(events),
    .groups = "drop"
  )  %>%
  ungroup() %>%
  group_by(tx, prevalence) %>%
  mutate(events_rate = events / max(events),
         surv_prop = (1-surv)/max(1-surv)) %>%
  ungroup()

colors <- c("0" = "#37293F", "1" = "#9f84af")

linetypes <- c(
  "75%" = "twodash",
  "50%" = "solid",
  "25%" = "dashed",
  "10%" = "dotted",
  "5%" = "dotdash"
)

p_time <- ggplot(df_summary_rob, aes(x = time, y = n.eff,
                         color = tx,
                         linetype = prevalence,
                         group = interaction(tx, prevalence))) +
  geom_hline(yintercept = 1000, linetype = "dotted", color = "grey50") +
  geom_hline(yintercept = 500,  linetype = "dotted", color = "grey50") +
  geom_line(linewidth = 1) +
  scale_color_manual(values = colors, name = "Group") +
  scale_linetype_manual(values = linetypes, name = "Baseline event rate") +
  labs(x = "Time", y = "Effective sample size") +
  scale_x_continuous(limits = c(0, 10)) +
  scale_y_continuous(limits = c(0, 1600)) +
  theme_minimal()

p_event <- ggplot(df_summary_rob, aes(x = events, y = n.eff,
                           color = tx,
                           linetype = prevalence,
                           group = interaction(tx, prevalence))) +
  geom_line(linewidth = 1) +
  geom_hline(yintercept = 1000, linetype = "dotted", color = "grey50") +
  geom_hline(yintercept = 500,  linetype = "dotted", color = "grey50") +
  scale_color_manual(values = colors, name = "Group") +
  scale_linetype_manual(values = linetypes, name = "Baseline event rate") +
  labs(x = "Number of events") +
  scale_x_continuous(limits = c(0, 1000)) +
  scale_y_continuous(limits = c(0, 1600)) +
  theme_minimal() +
  theme( axis.text.y = element_blank(), 
         axis.ticks.y = element_blank(),
         axis.title.y = element_blank())


p_event_prop <- ggplot(df_summary_rob, aes(x = events_rate, y = n.eff,
                           color = tx,
                           linetype = prevalence,
                           group = interaction(tx, prevalence))) +
  geom_line(linewidth = 1) +
  geom_hline(yintercept = 1000, linetype = "dotted", color = "grey50") +
  geom_hline(yintercept = 500,  linetype = "dotted", color = "grey50") +
  scale_color_manual(values = colors, name = "Group") +
  scale_linetype_manual(values = linetypes, name = "Baseline event rate") +
  labs(x = "Proportion of total events") +
  scale_x_continuous(limits = c(0, 1)) +
  scale_y_continuous(limits = c(0, 1600)) +
  theme_minimal() +
  theme( axis.text.y = element_blank(), 
         axis.ticks.y = element_blank(),
         axis.title.y = element_blank()) 

# Combine the plots using patchwork and add a single legend on the right
combined_plot <- ( ( p_time | p_event | p_event_prop )  ) +
  plot_layout(guides = "collect") +
  plot_annotation(
    title = "Effective sample size for different event rates",
    theme = theme(plot.title = element_text(size = 14, face = "bold", hjust = 0))
  ) +
  theme(legend.position = "right")

# Print the combined plot
print(combined_plot)
```

```{r}
# Seed for consistency
set.seed(1923)

shape = c(0.5) # shape 1 means exponential
# make a vector of all lambda's from 0.01 to 0.99 in the scale parameter, so do lambda(x, shape) for all x in that range
scale =  sapply(seq(0.01, 0.99, by = 0.01), function(x) lambda(x, shape)) # scale
maxtime = 10 # maxtime
hr = 2 # hazard ratio
n = 1000 # number per group
time_grid <- seq(0, 10, by = 0.001)
n_sim <- 10  # choose number of repeats

df <- bind_rows(
  lapply(1:n_sim, function(r)
    run_one(r, scale, shape, maxtime, hr, n, time_grid, model_type = "standard")
  )
)

# Levels need to be from 1% to 99% in the prevalence variable as character
#df$prevalence <- factor(df$prevalence, levels = seq(0.01, 0.99, 0.01) )
df <- df %>%
  group_by(tx, prevalence) %>%
  mutate(events = max(n.risk) - n.risk)
df_summary <- df %>%
  group_by(time, tx, prevalence) %>%
  summarise(
    n.eff = mean(n.eff),
    surv = mean(surv),
    events = mean(events),
    .groups = "drop"
  )  %>%
  ungroup() %>%
  group_by(tx, prevalence) %>%
  mutate(events_rate = events / max(events),
         surv_prop = (1-surv)/max(1-surv)) %>%
  ungroup()

max_points <- df_summary %>%
  group_by(tx, prevalence) %>%
  slice_max(order_by = 1-surv, n = 1) %>%
  ungroup()

p_surv <- ggplot(df_summary[df_summary$tx == 0,], aes(x = 1-surv, y = n.eff,
                         color = prevalence,
                         group = interaction(tx, prevalence))) +
  geom_line(linewidth = 0.7, alpha = 0.3) +
  geom_line(data = max_points,
            aes(x = 1-surv, y = n.eff, group = 1),
            color = "#37293F", linewidth = 1.2, linetype = "solid") +
  scale_color_gradient(low="#2E7691", high = "#C2666B") +
  scale_linetype_manual(values = linetypes, name = "Baseline event rate") +
  labs(x = "Event incidence", y = "Effective sample size") +
    scale_x_continuous(limits = c(0, 1)) +
  scale_y_continuous(limits = c(0, 1600)) +
  theme_minimal()

library(ggnewscale) # For multiple color scales

p_surv <- ggplot(df_summary, aes(x = 1-surv, y = n.eff)) +
  # For tx == 0
  geom_line(data = subset(df_summary, tx == 0),
            aes(color = prevalence, group = interaction(tx, prevalence)),
            linewidth = 0.5, alpha = 0.2) +
  scale_color_gradient(low = "#a8d3e3", high = "#153744", name = "Prevalence (tx=0)") +
  # For tx == 1
  new_scale_color() +
  geom_line(data = subset(df_summary, tx == 1),
            aes(color = prevalence, group = interaction(tx, prevalence)),
            linewidth = 1, alpha = 0.3) +
  scale_color_gradient(low = "#dfafb2", high = "#8b373c", name = "Prevalence (tx=1)") +
  # Add solid line for max points (new color scale)
  new_scale_color() +
  geom_line(data = max_points,
            aes(x = 1-surv, y = n.eff, group = tx, color = factor(tx)),
            linewidth = 1, linetype = "solid") +
  scale_color_manual(values = c("0" = "#2E7691", "1" = "#C2666B"),
                     name = "Treatment") +
  labs(x = "Event incidence", y = "Effective sample size") +
  scale_x_continuous(limits = c(0, 1)) +
  scale_y_continuous(limits = c(0, 1600)) +
  theme_minimal()

```


<!-- events_50 <- simsurv( dist="weibull",  -->
<!--                    gammas=shape,  -->
<!--                    lambdas=scale[1],  -->
<!--                    x=data.frame(tx = c(rep(0, n/2), rep(1, n/2))), -->
<!--                    betas = c("tx" = log(hr)), -->
<!--                    maxt=maxtime) -->
<!-- events_25 <- simsurv( dist="weibull",  -->
<!--                    gammas=shape,  -->
<!--                    lambdas=scale[2],  -->
<!--                    x=data.frame(tx = c(rep(0, n/2), rep(1, n/2))), -->
<!--                    betas = c("tx" = log(hr)), -->
<!--                    maxt=maxtime) -->
<!-- events_10 <- simsurv( dist="weibull",  -->
<!--                    gammas=shape,  -->
<!--                    lambdas=scale[3],  -->
<!--                    x=data.frame(tx = c(rep(0, n/2), rep(1, n/2))), -->
<!--                    betas = c("tx" = log(hr)), -->
<!--                    maxt=maxtime) -->
<!-- events_5 <- simsurv( dist="weibull",  -->
<!--                    gammas=shape,  -->
<!--                    lambdas=scale[4],  -->
<!--                    x=data.frame(tx = c(rep(0, n/2), rep(1, n/2))), -->
<!--                    betas = c("tx" = log(hr)), -->
<!--                    maxt=maxtime) -->
<!-- events_50$tx = c(rep(0, n/2), rep(1, n/2)) -->
<!-- events_25$tx = c(rep(0, n/2), rep(1, n/2)) -->
<!-- events_10$tx = c(rep(0, n/2), rep(1, n/2)) -->
<!-- events_5$tx = c(rep(0, n/2), rep(1, n/2)) -->

<!-- cox_50 <- coxph(Surv(eventtime, status) ~ tx, data = events_50) -->
<!-- cox_25 <- coxph(Surv(eventtime, status) ~ tx, data = events_25) -->
<!-- cox_10 <- coxph(Surv(eventtime, status) ~ tx, data = events_10) -->
<!-- cox_5 <- coxph(Surv(eventtime, status) ~ tx, data = events_5) -->

<!-- cox_50_n0 <- survfit_n(survfit(cox_50, newdata = data.frame(tx = 0)), cox_50, coef=T, chaz=T) -->
<!-- cox_25_n0 <- survfit_n(survfit(cox_25, newdata = data.frame(tx = 0)), cox_25, coef=T, chaz=T) -->
<!-- cox_10_n0 <- survfit_n(survfit(cox_10, newdata = data.frame(tx = 0)), cox_10, coef=T, chaz=T) -->
<!-- cox_5_n0 <- survfit_n(survfit(cox_5, newdata = data.frame(tx = 0)), cox_5, coef=T, chaz=T) -->

<!-- cox_50_n1 <- survfit_n(survfit(cox_50, newdata = data.frame(tx = 1)), cox_50, coef=T, chaz=T) -->
<!-- cox_25_n1 <- survfit_n(survfit(cox_25, newdata = data.frame(tx = 1)), cox_25, coef=T, chaz=T) -->
<!-- cox_10_n1 <- survfit_n(survfit(cox_10, newdata = data.frame(tx = 1)), cox_10, coef=T, chaz=T) -->
<!-- cox_5_n1 <- survfit_n(survfit(cox_5, newdata = data.frame(tx = 1)), cox_5, coef=T, chaz=T) -->

<!-- cox_50_robust <- coxph(Surv(eventtime, status) ~ tx, data = events_50, robust = TRUE) -->
<!-- cox_25_robust <- coxph(Surv(eventtime, status) ~ tx, data = events_25, robust = TRUE) -->
<!-- cox_10_robust <- coxph(Surv(eventtime, status) ~ tx, data = events_10, robust = TRUE) -->
<!-- cox_5_robust <- coxph(Surv(eventtime, status) ~ tx, data = events_5, robust = TRUE) -->

<!-- cox_50_robust_n0 <- survfit_n(survfit(cox_50_robust, newdata = data.frame(tx = 0)), cox_50_robust, coef=T, chaz=T) -->
<!-- cox_25_robust_n0 <- survfit_n(survfit(cox_25_robust, newdata = data.frame(tx = 0)), cox_25_robust, coef=T, chaz=T) -->
<!-- cox_10_robust_n0 <- survfit_n(survfit(cox_10_robust, newdata = data.frame(tx = 0)), cox_10_robust, coef=T, chaz=T) -->
<!-- cox_5_robust_n0 <- survfit_n(survfit(cox_5_robust, newdata = data.frame(tx = 0)), cox_5_robust, coef=T, chaz=T) -->

<!-- cox_50_robust_n1 <- survfit_n(survfit(cox_50_robust, newdata = data.frame(tx = 1)), cox_50_robust, coef=T, chaz=T) -->
<!-- cox_25_robust_n1 <- survfit_n(survfit(cox_25_robust, newdata = data.frame(tx = 1)), cox_25_robust, coef=T, chaz=T) -->
<!-- cox_10_robust_n1 <- survfit_n(survfit(cox_10_robust, newdata = data.frame(tx = 1)), cox_10_robust, coef=T, chaz=T) -->
<!-- cox_5_robust_n1 <- survfit_n(survfit(cox_5_robust, newdata = data.frame(tx = 1)), cox_5_robust, coef=T, chaz=T) -->

<!-- df <- bind_rows( -->
<!--   extract_sf(cox_50_n0, 0, "50%"), -->
<!--   extract_sf(cox_50_n1, 1, "50%"), -->
<!--   extract_sf(cox_25_n0, 0, "75%"), -->
<!--   extract_sf(cox_25_n1, 1, "75%"), -->
<!--   extract_sf(cox_10_n0, 0, "90%"), -->
<!--   extract_sf(cox_10_n1, 1, "90%"), -->
<!--   extract_sf(cox_5_n0, 0, "95%"), -->
<!--   extract_sf(cox_5_n1, 1, "95%") -->
<!-- ) -->


## References {.unnumbered}

:::{#refs}

:::